---
title: "Simulation results"
author: "Emmanuelle Comets"
date: '2022-08-18'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Folders
workDir<-getwd() 

# @Eco
workDir<-"/home/eco/work/saemix/saemixextension/paperSaemix3"
simDir <- "/home/eco/work/saemix/discreteEval"
saemixDir <- "/home/eco/work/saemix/saemixextension"
setwd(workDir)

# Libraries
library(ggplot2)

saveFigs<-TRUE

```


## Content



## Binary model

Simulations run using scripts (run before, only the results are used here)

```{r binaryModelSimulations, echo=FALSE}
if(FALSE) {
  setwd(simDir)
  system("R CMD BATCH eval_binaryOrig.R eval_binaryOrig.out")  
  system("R CMD BATCH eval_binaryIIV.R eval_binaryIIV.out")  
  
}

```

Results

- two scenarios, with design of the toenail study (per protocol)
  - IIV on $\theta_1$ only: bias less than 5\% for all parameters (mean 4\% but median is 8.5\% ?)
  - IIV on both parameters with 50\% IIV: some bias on $\theta_2$ and $\beta$ (~10\%) and a little on $\omega_{\theta_2}$ (6.5\%)
    - the bias on $\omega_{\theta_2}$ disappears when we start from the true parameters but the bias on $\theta_2$ and $\beta$ remains around 10\%
- comparing settings for initial values
  - no difference w/r starting values of the parameters $\Rightarrow$ runs seem to be stable with 10 chains

- Expected RSE from PFIM
  - binaryOrig: RSE($\theta_1$)=-24.86142 RSE($\theta_2$)=-11.63633 RSE($\beta$)=-38.29206  RSE($\omega_1^2$)=20.39756 
    - all parameters should be well estimated
  - binaryIIV:  RSE($\theta_1$)=-10.70774  RSE($\theta_2$)=-20.99451 RSE($\beta$)=-112.04915   RSE($\omega_1^2$)=36.80376  RSE($\omega_2^2$)= 65.15504 
    - $\beta$ expected to be poorly estimated (but $\theta_2$ should be well estimated)
  - in both cases, good agreement between predicted SE and empirical SE (obtained as SD(estimates) over the 200 simulations) for all parameters (with sometimes empirical SE smaller than predicted SE, for $\theta_1$ in original scenario and for $\omega_{\theta_2}$ in scenario with two IIV)
    - compare the SD computed here (=empirical SE) with predicted SE from seFIMAGQ_true.res in the result folder
    - slightly smaller for $\omega_{\theta_1}$ and smaller by 50\% $\omega_{\theta_2}$in IIV and for $\theta_1$ in original scenari, otherwise very close
    
- Compared one run with Monolix and very similar estimates when using 10 chains and the same CI


```{r binaryModelPerformance, echo=FALSE}
scenario <- "binaryOrig"

par.fix<-c(-1.71, -0.39, -0.15)
par.om <- c(4.02) 
param<-c(par.fix, par.om)
npar<-length(param)

ytab<-NULL
for (setting in c("true","pop","far")) {
  res1<-read.table(file.path(simDir, scenario,"results",paste0(setting,"fit_",scenario,".tab")), header=T)
  for(icol in 1:npar) {
  y1<-data.frame(parameter=colnames(res1)[icol+1], value=(res1[,(icol+1)]-param[icol])/param[icol]*100 ,scenario=scenario, setting=setting)
  ytab<-rbind(ytab, y1)
  }
}

plotOrig <- ggplot(data=ytab, aes(x=parameter, y=value, fill=setting)) + geom_boxplot() + geom_hline(yintercept=c(-5,5), linetype="dotted") + geom_hline(yintercept=c(-10,10), linetype="dashed") + xlab("Parameter") + ylab("Relative estimation error (%)")
print(plotOrig)

if(saveFigs) {
  namfig<-file.path(simDir, scenario,"results",paste0("ree_",scenario,".eps"))
  cairo_ps(file = namfig, onefile = TRUE, fallback_resolution = 600, height=8.27, width=11.69)
  plot(plotOrig)
  dev.off()
}

truepar <- data.frame(parameter=colnames(res1)[2:(1+npar)], value=param)

setting<-"pop"
res1<-read.table(file.path(simDir, scenario,"results",paste0(setting,"fit_",scenario,".tab")), header=T)
for(icol in 1:4) {
  print(summary(res1[,(icol+1)]-param[icol]))
  print(summary((res1[,(icol+1)]-param[icol])/param[icol]))
  if(icol==4) print(sd(res1[,(icol+1)]**2)) else print(sd(res1[,(icol+1)]))
}

### two IIV around 50%

scenario <- "binaryIIV"
par.fix<-c(-1.71, -0.39, -0.15)
par.om <- c(1, 0.2) # around 50% variability for both
param<-c(par.fix, par.om)
npar<-length(param)

ytab<-NULL
for (setting in c("true","pop","far")) {
  res1<-read.table(file.path(simDir, scenario,"results",paste0(setting,"fit_",scenario,".tab")), header=T)
  for(icol in 1:npar) {
  y1<-data.frame(parameter=colnames(res1)[icol+1], value=(res1[,(icol+1)]-param[icol])/param[icol]*100 ,scenario=scenario, setting=setting)
  ytab<-rbind(ytab, y1)
  }
}

plotIIV <- ggplot(data=ytab, aes(x=parameter, y=value, fill=setting)) + geom_boxplot() + geom_hline(yintercept=c(-5,5), linetype="dotted") + geom_hline(yintercept=c(-10,10), linetype="dashed") + xlab("Parameter") + ylab("Relative estimation error (%)")
print(plotIIV)

if(saveFigs) {
  namfig<-file.path(simDir, scenario,"results",paste0("ree_",scenario,".eps"))
  cairo_ps(file = namfig, onefile = TRUE, fallback_resolution = 600, height=8.27, width=11.69)
  plot(plotIIV)
  dev.off()
}

res1<-read.table(file.path(simDir, scenario,"results",paste0("popfit_",scenario,".tab")), header=T)
for(icol in 1:length(param)) {
  print(summary(res1[,(icol+1)]-param[icol]))
  print(summary((res1[,(icol+1)]-param[icol])/param[icol]))
  if(icol>3) print(sd(res1[,(icol+1)]**2)) else print(sd(res1[,(icol+1)]))
}


```

