---
title: "Saemix 3 - time-to-event models"
author: "Emmanuelle"
date: "09/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


/home/eco/biblio/survival/vanwijk22_CPTPSP_choosingHazardFctTTETutorialShinyApp.pdf

## Version
Use saemix version $\geq$ 3.2

- correction of a bug in the simulation file
- added a simulation function for non Gaussian data models
  - **TODO** test if function is NULL !!!

## Objective

Run single and repeated time-to-event models in **saemix**

This notebook uses additional code from the **saemix** development github, not yet integrated in the package. The *workDir* folder in the next chunk of code points to the folder where the user stored this code, and is needed to run the notebook (*workDir* defaults to the current working directory). Specifically, the notebook loads:

- code for different bootstraps in non-linear mixed effect models (Comets et al. 2021 and submitted)
  - the bootstrap runs have been performed previously and are stored in files to be read
    - bootstraps can be run instead by switching the *runBootstrap* variable to TRUE in the first chunk of code
    - in the code, the number of bootstraps is set to 10 for speed but we recommend to use at least 200 for a 90\% CI.
  - this can be changed in the following change of code by uncommenting the line *nboot<-200* and setting the number of bootstrap samples (this may cause memory issues in **Rstudio** with older machines, if this is the case we recommend executing the code in a separate script)
- code for the MC/AGQ provided by Sebastian Ueckert (Ueckert et al. 2017)
  - again if memory issues arise the code can be run in a separate script.

The current notebook can be executed to create an HMTL or PDF output with comments and explanations. A script version containing only the R code is also given as *saemix3_tteModel.R* in the same folder.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Folders
workDir<-getwd() 

# @Eco
workDir<-"/home/eco/work/saemix/saemixextension/paperSaemix3"
saemixDir <- "/home/eco/work/saemix/saemixextension"
setwd(workDir)

# Libraries
library(saemix)

# Libraries needed to compute the FIM by AGQ
library(R6)
library(pracma)
library(compiler)
library(statmod)
library(Matrix)
library(randtoolbox)

# FIM by MC/AGQ (code S. Ueckert)

dirAGQ<-file.path(saemixDir,"fimAGQ")

# Bootstrap code
source(file.path(saemixDir, "bootstrap", "saemix_bootstrap.R"))

# Code to compute the exact FIM by MC/AGQ

# library(ggplot2)
# library(MASS)
# library(rlang)
# library(gridExtra)
library(tidyverse)

# Whether to save the plots
saveFigs<-FALSE
figDir <- getwd()

# Number of bootstrap samples
runBootstrap <- FALSE # to read the results from disk
nboot <-10
# nboot <- 200

```

### Time-to-event model for a single event

#### Data description - Alexandra change here or later for competitive risk ?

Lung data from the survival package **[copy here from examplesDocumentation.R]**
- reformatting changes
  - saemix format: added time=0
  - created one column for status (dead or alive, recoded as 1/0) and one for censoring (0/1)
  - removed subjects with missing age, institution, sex, or ph scores (ecog and karno)

**Checks**

- The `Surv` function from the `survival` package creates a survival object for use as the response in a model formula.
  - one entry for each subject that is the survival time, which is followed by a `+` if the subject was censored
  - transform lung.saemix in the Surv format to check the survival function w/r saemix fit
- Weibull model
  - parametric survival function $$ S(t) = e^{ - \left( \frac{t}{\lambda} \right) ^{\beta}}$$
- Also tried computing a SE for $S(t)$ using the delta-method where the vector of derivatives for the survival function of Weibull model are:

$$ \begin{pmatrix}
\frac{\delta S}{\delta \lambda} \\
\frac{\delta S}{\delta \beta}
\end{pmatrix} = 
\begin{pmatrix}
\frac{\beta}{\lambda} \; \left( \frac{t}{\lambda} \right)^{\beta} e^{-  \left( \frac{t}{\lambda} \right)^{\beta}} \\ 
- \ln \left( \frac{t}{\lambda} \right) \; \left( \frac{t}{\lambda} \right)^{\beta} e^{-  \left( \frac{t}{\lambda} \right)^{\beta}} \\ 
\end{pmatrix}$$

  - works pretty well compared to the non-parametric KM estimate
#### Exploring data

**TODO** Kaplan-Meier plot, other plots, maybe

The plot is created using **ggplot2**.

#### Statistical model

A nice review of the more frequent hazard functions used in parametric models of TTE data has recently been van Wijk and Simonsson (*CPT:PSP* 2022), including a Shiny app to explore their shape and how to set initial parameters. These models are very sensitive to the initial parameter estimates and their variance, therefore using 

## References

**Comets E**, Rodrigues C, Jullien V, Ursino M (2021). Conditional non-parametric bootstrap for non-linear mixed effect models. *Pharmaceutical Research*, 38: 1057-66.

**Ueckert S**, Mentré F (2017). A new method for evaluation of the Fisher information matrix for discrete mixed effect models using Monte Carlo sampling and adaptive Gaussian quadrature. *Computational Statistics and Data Analysis*, 111: 203-19. \url{10.1016/j.csda.2016.10.011}

**van Wijk R**, Simonsson U (2022). Finding the right hazard function for time- to-event modeling: A tutorial and Shiny application. *Clinical Pharmacokinetics and Therapeutics: Pharmacometrics and Systems Pharmacology*

