---
title: "Use case for bootstrap methods in saemix"
author: "Emmanuelle Comets"
date: "17/09/2020"
output:
  pdf_document: default
  html_document: default
---

## Contents

The present document is an R markdown notebook designed to showcase the different bootstrap methods in the paper **Conditional non-parametric bootstrap for non-linear mixed effect models** submitted by E. Comets et al (currently submitted to **Biostatistics**, September 2020). This notebook uses code from the development version of the *saemix* package, available on github: <https://github.com/saemixdevelopment/saemixextension>. An easy way to ensure all the functions are available is to clone the project using Rstudio, and follow updates by pull requests. Folders are relative to the installation folder, assuming the same architecture as on the github repository.

In addition, several libraries need to be installed to run the present code:

- ggplot2 and its dependencies
- MASS

The full set of simulated data and bootstrap distributions used for the paper have been uploaded to the Zenodo archive system <https://doi.org/10.5281/zenodo.4059718>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## If required, set your working directory, otherwise will run in the current directory
## workDir<-"/my/path/to/saemixextension-master"
workDir<-"."

workDir<-"/home/eco/work/saemix/saemixextension"
knitr::opts_chunk$set(root.dir = workDir)
progDir<-file.path(workDir,"R")
```

### Loading functions

As we use the development version of *saemix*, this chunk loads the functions from the package, as well as necessary libraries.

```{r loadLibFct}
# Libraries
library(ggplot2)
library(MASS)

# Sourcing saemix functions
{
  source(file.path(progDir,"aaa_generics.R"))
  #source(file.path(progDir,"global.R"))
  source(file.path(progDir,"SaemixData.R"))
  source(file.path(progDir,"SaemixRes.R"))
  source(file.path(progDir,"SaemixModel.R"))
  source(file.path(progDir,"SaemixObject.R"))
  source(file.path(progDir,"main.R"))
  source(file.path(progDir,"func_aux.R"))
  source(file.path(progDir,"main_initialiseMainAlgo.R"))
  source(file.path(progDir,"main_estep.R"))
  source(file.path(progDir,"main_mstep.R"))
  source(file.path(progDir,"func_FIM.R"))
  source(file.path(progDir,"func_plots.R"))
  source(file.path(progDir,"func_distcond.R"))
  source(file.path(progDir,"func_simulations.R"))
  source(file.path(progDir,"compute_LL.R"))
  source(file.path(progDir,"func_estimParam.R"))
}

# Bootstrap functions
source(file.path(workDir,"bootstrap","saemix_bootstrap.R"))
```

## Bootstrap distributions for one simulated dataset

### Plot of simulated data for one simulation in each scenario

Below, we load the simulated data for the 6 scenarios presented in the paper (we randomly sampled one of the 200 simulations used to evaluate the performances of the different bootstrap approaches, simulation 87) and plot it using the *ggplot2* package (Supplemental material Figure 12).

```{r simData87, echo=FALSE}
datDir<-file.path(workDir,"bootstrap","data")
isim<-87

ypd<-NULL
for(myloop in 1:6) {
  if(myloop==1) {namscen<-"pdemax.rich";iscenar<-"Hill2";namsimdat<-"data_pdemax";name.scenario<-"SEmax,R"}
  if(myloop==2) {namscen<-"pdemax.sparse";iscenar<-"Hill2";namsimdat<-"data_pdemax";name.scenario<-"SEmax,S"}
  if(myloop==3) {namscen<-"pdhillhigh.rich";iscenar<-"Hill3";namsimdat<-"data_pdhillhigh";name.scenario<-"SHill,R"}
  if(myloop==4) {namscen<-"pdhillhigh.sparse";iscenar<-"Hill4";namsimdat<-"data_pdhillhigh";name.scenario<-"SHill,S"}
  if(myloop==5) {namscen<-"pdhillhigh.rich";iscenar<-"Hill12";namsimdat<-"data_pdhillhigh";name.scenario<-"SHill,smallR"}
  if(myloop==6) {namscen<-"pdhillhigh.sparse";iscenar<-"Hill11";namsimdat<-"data_pdhillhigh";name.scenario<-"SHill,smallS"}
  namfile<-file.path(datDir,paste("data_",namscen,"_sim",isim,".tab",sep=""))
  xdat<-read.csv(namfile,header=T,sep=" ")
  if(myloop>4) {
    if(myloop==5) nsuj<-20 else nsuj<-80
    zesuj<-unique(xdat[,1])
    ifirst<-rep(1:(nsuj/4),4)+rep(c(0:3)*length(zesuj)/4,each=(nsuj/4)) # First nsuj/4 subjects in each group
    xdat<-xdat[xdat$id %in% ifirst,]
  }
  ypd<-rbind(ypd,cbind(xdat,scenario=name.scenario))
}
  
plot.data<-ggplot(data=ypd,aes(x=dose,y=y, group=id)) + geom_point(col="red4",alpha=0.6)+geom_line(col="red4",alpha=0.6) + 
    xlab("Dose") + ylab("Response") + theme_bw() +
    theme(axis.title.x = element_blank(),axis.text.x = element_text(size=9, angle=30, hjust=1), legend.position = "none") + 
    facet_wrap(~scenario, ncol=2)

print(plot.data)
```

### Fitting the simulated data in replication 87, scenario SHill,R, with saemix

```{r saemixFit87.rich, echo=FALSE, warning=FALSE}
isim<-87
nboot<-100
namscen<-"pdhillhigh.rich"
iscenar<-"Hill3"

mod.par<-c(E0=5,Emax=30,ED50=500,gamma=3)
mod.iiv<-c(E0=0.09,Emax=0.49,ED50=0.49,gamma=0)
omega<-diag(mod.iiv)
omega[3,2]<-omega[2,3]<-0.245
sigm<-0.1
modelhill<-function(psi,id,xidep) {
  # input:
  #   psi : matrix of parameters (4 columns, E0, Emax, E50, gamma)
  #   id : vector of indices 
  #   xidep : dependent variables (same nb of rows as length of id)
  # returns:
  #   a vector of predictions of length equal to length of id
  dose<-xidep[,1]
  e0<-psi[id,1]
  emax<-psi[id,2]
  e50<-psi[id,3]
  gamma<-psi[id,4]
  f<-e0+emax*dose**gamma/(e50**gamma+dose**gamma)
  return(f)
}

namfile<-file.path(datDir,paste("data_",namscen,"_sim",isim,".tab",sep=""))
saemix.data<-saemixData(name.data=namfile, header=T,
                        name.group=c("id"),name.predictors=c("time","amt"),
                        name.response=c("conc"),name.covariates=NULL,
                        units=list(x="hr",y="mg/L",covariates=c()))
saemix.model<-saemixModel(model=modelhill,description="Hill model", 
                          psi0=matrix(mod.par,ncol=4, byrow=TRUE,dimnames=list(NULL, c("E0","Emax","ED50","Gamma"))),transform.par=c(1,1,1,1),
                          covariance.model=matrix(c(1,0,0,0,0,1,1,0,0,1,1,0,0,0,0,0),ncol=4,byrow=TRUE),omega.init = omega,error.model="proportional", error.init = c(0,sigm))

saemix.options<-list(fix.seed=F,directory="current",displayProgress=FALSE, save.graphs=FALSE,print=FALSE)
saemix.fit<-saemix(saemix.model,saemix.data,saemix.options)
 print(saemix.fit@results)

om.estim<-c(diag(saemix.fit@results@omega)[1:3], saemix.fit@results@omega[2,3], saemix.fit@results@respar[2])
par.estim<-c(saemix.fit@results@fixed.effects,om.estim)
```

### Bootstrap distributions for the simulated data in replication 87, scenario SHill,R

The following code loads the bootstrap distributions obtained by the four tested bootstraps for the simulated data in replication 87 (shown the first plot of the second line in the previous figure) from the result files, and plots these for each parameter. This corresponds to Figure 1 in the paper (estimated values may slightly vary due to the stochastic nature of the SAEM algorithm).

Note: 500 bootstrap samples were generated for each of the first 4 scenarios, but in the paper we harmonise across the different scenarios to 100 samples through the *nboot<-100* command in the code below.

```{r bootstrapDistributions87.rich, echo=FALSE}
resDir<-file.path(workDir,"bootstrap","results")
isim<-87
nboot<-100
namscen<-"pdhillhigh.rich"
iscenar<-"Hill3"
namsimdat<-"data_pdemax"

modpar2<-c(mod.par,mod.iiv[1:3],0.245,sigm)
ypd1<-NULL
for(typeboot in c("Case","NP","NPc","Par")) {
  namfile<-file.path(resDir,paste("scenario",iscenar,"_bootstrap",typeboot,"_sim",isim,".res",sep=""))
  resboot1<-read.table(namfile,header=T)
  resboot1<-resboot1[1:nboot,] 
  for(icol in 1:9) {
    ypd1<-rbind(ypd1,data.frame(rep=resboot1[,1],Param=colnames(resboot1)[(icol+1)],value=resboot1[,(icol+1)], Bootstrap=typeboot,stringsAsFactors=FALSE))
  }
}
nampar<-colnames(resboot1)[2:10]
ypd1$Param<-factor(ypd1$Param,levels = nampar)
ypd1$Bootstrap<-factor(ypd1$Bootstrap, levels=unique(ypd1$Bootstrap))
levels(ypd1$Bootstrap)[levels(ypd1$Bootstrap)=="NPc"]<-"cNP"
ypd1.fix<-ypd1[ypd1$Param %in% unique(ypd1$Param)[1:4],]
ypd1.iiv<-ypd1[ypd1$Param %in% unique(ypd1$Param)[5:9],]

df<-data.frame(Param=factor(nampar), value=modpar2, value2=par.estim) 

plot.density<-ggplot(data=ypd1) + geom_density(aes(value,fill=Bootstrap), alpha=0.5) + 
  geom_vline(data=df,aes(xintercept=value),colour="red",size=1.2) + geom_vline(data=df,aes(xintercept=value2),colour="blue",size=1.2) +
  theme_bw() + theme(axis.title.x = element_blank(),axis.text.x = element_text(size=9, angle=30, hjust=1), legend.position = "top") + 
  scale_fill_manual(values = c("red4", "darkblue", "steelblue1","seagreen4","snow4")) +  facet_wrap(~Param, ncol=3, scales = 'free')

print(plot.density)

# Saving to a PDF file (requires cairo)
cairo_pdf(file.path(workDir,"bootstrap","fig1_paper.pdf"),width = 11, height=8)
print(plot.density)
dev.off()

```

## Running bootstrap methods on one dataset (replication 87, scenario SHill,R)

In the following we estimate bootstrap distributions with 100 samples using the different bootstrap methods through the *saemix.bootstrap()* function, and recreate the previous figure. Due to stochastic variability the bootstrap distributions are different from before but their shape and amplitude is similar.

**Warning:** this code may take some time to execute. My Rstudio tends to run out of memory and crash before rendering, so I generated the pdf using the command *rmarkdown::render("./comets_condBoostrapSaemix.Rmd")* in a plain R window.


```{r bootstrapRuns87, echo=FALSE, warning=FALSE, results='hide'}
saemix.bootOpt<-list(fix.seed=F,directory="current",displayProgress=F, save.graphs=F, map=F, ll.is=F, print=FALSE)
nboot<-100

start_time <- Sys.time()
boot.case<-saemix.bootstrap(saemix.fit, nboot=nboot, method="case")
boot.cNP<-saemix.bootstrap(saemix.fit, nboot=nboot, method="conditional")
boot.NP<-saemix.bootstrap(saemix.fit, nboot=nboot, method="residual")
boot.Par<-saemix.bootstrap(saemix.fit, nboot=nboot, method="parametric")
end_time <- Sys.time()
```

```{r bootstrapRuns87.res, echo=FALSE, warning=FALSE}
# Summarising the results
res<-data.frame(Case=colMeans(boot.case[,2:dim(boot.case)[2]]), NP=colMeans(boot.NP[,2:dim(boot.case)[2]]),cNP=colMeans(boot.cNP[,2:dim(boot.case)[2]]), Par=colMeans(boot.Par[,2:dim(boot.case)[2]]))
sdres<-data.frame(Case=apply(boot.case[,2:dim(boot.case)[2]],2,sd), NP=apply(boot.NP[,2:dim(boot.case)[2]],2,sd), cNP=apply(boot.cNP[,2:dim(boot.case)[2]],2,sd), Par=apply(boot.Par[,2:dim(boot.case)[2]],2,sd))

boot.est<-cbind(True=df$value,res[1:dim(df)[1],])
for(icol in 2:5) {
  boot.est[,icol]<-format(boot.est[,icol],nsmall=2, digits=1, scientific=FALSE)
  sdres[,(icol-1)]<-format(sdres[,(icol-1)],nsmall=1, digits=1, scientific=FALSE)
  boot.est[,icol]<-paste(boot.est[,icol]," (",sdres[1:dim(df)[1],(icol-1)],")",sep="")
}
cat("Bootstrap estimates (SD)\n")
print(boot.est)
cat("Run time:",end_time - start_time,"\n")

# Recreating previous plot
ypd2<-NULL
for(typeboot in c("Case","NP","NPc","Par")) {
  if(typeboot=="Case") resboot1<-boot.case
  if(typeboot=="NP") resboot1<-boot.NP
  if(typeboot=="NPc") resboot1<-boot.cNP
  if(typeboot=="Par") resboot1<-boot.Par
  colnames(resboot1)[2:10]<-nampar
  for(icol in 1:9) {
    ypd2<-rbind(ypd2,data.frame(rep=resboot1[,1],Param=colnames(resboot1)[(icol+1)],value=resboot1[,(icol+1)], Bootstrap=typeboot, stringsAsFactors=FALSE))
  }
}
ypd2$Param<-factor(ypd2$Param, levels = nampar)
ypd2$Bootstrap<-factor(ypd2$Bootstrap, levels=unique(ypd2$Bootstrap))
levels(ypd2$Bootstrap)[levels(ypd2$Bootstrap)=="NPc"]<-"cNP"
ypd2.fix<-ypd2[ypd2$Param %in% unique(ypd2$Param)[1:4],]
ypd2.iiv<-ypd2[ypd2$Param %in% unique(ypd2$Param)[5:9],]

plot.density2<-ggplot(data=ypd2) + geom_density(aes(value,fill=Bootstrap), alpha=0.5) + 
  geom_vline(data=df,aes(xintercept=value),colour="red",size=1.2) + geom_vline(data=df,aes(xintercept=value2),colour="blue",size=1.2) +
  theme_bw() + theme(axis.title.x = element_blank(),axis.text.x = element_text(size=9, angle=30, hjust=1), legend.position = "top") + 
  scale_fill_manual(values = c("red4", "darkblue", "steelblue1","seagreen4","snow4")) +  facet_wrap(~Param, ncol=3, scales = 'free')

print(plot.density2)

cairo_pdf(file.path(workDir,"bootstrap","fig1_interactive.pdf"),width = 11, height=8)
print(plot.density2)
dev.off()

```
