---
#title: "Using saemix package to fit joint models"
#author: "Alexandra Lavalley-Morelle"
#date: "27/09/2023"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This Rmarkdown document helps user to buid and fit joint models using the code extension available on Github.
This document relies on the examples presented in the article. 

## Prothrombin example

We consider the data available in package JM: the prothro data set. 488 liver cirrhosis patients are followed at most 12 days
with prothrombin measurements. Status (dead/censored) at the end of the follow-up is available for each patient. The objective is to assess the link between the individual prothrombin evolution with the risk of death.  

```{r Packages loading}
library(JM)
library(pracma)
library(ggplot2)
library(Cairo)
library(viridis)
library(rlang)
```

```{r Data loading}
data("prothro")
data("prothros")

gp = ggplot(data=prothro[which(prothro$id %in% 1:20),], aes(x=time, y=pro, group = id))+
  geom_point(lwd=1.5)+geom_line(col="#CC0033",lwd=0.8)+theme_classic()+
  ylab("Prothrombin observations")+xlab("Days")+
  theme(axis.text = element_text(size=14),axis.title = element_text(size=16))+
  ggtitle(label = "Prothrombin evolution for the first 20 patients of the data")
gp

table(prothros$death)
```

## Joint model fit using new saemix.multi() function
The extended code uses the same main functions as saemix package does. We therefore refer the user to the saemix documentation previously published for the detail of each function (see Comets et al. JSS, 2017). Briefly, the main function is the saemix.multi() function used to estimate the population parameters of the (joint) model. This function requires two mandatory arguments referring to (1) the model (saemixModel object) and the data (saemixData object). The third argument is optional and concerns the algorithm settings. 

#### Loading functions from Github

```{r loading functions}
saemixDir <- "C:/Users/AlexandraLAVALLEY/Documents/GitHub/saemixextension"
workDir <- file.path(saemixDir, "joint")
progDir<-file.path(saemixDir, "R")
setwd(workDir)

source(file.path(progDir,"aaa_generics.R"))
source(file.path(progDir,"SaemixData.R"))
source(file.path(progDir,"SaemixRes.R"))
source(file.path(progDir,"SaemixModel.R"))
source(file.path(progDir,"SaemixObject.R"))
source(file.path(progDir,"func_plots.R"))

source(file.path(workDir,"multi_aux.R"))
source(file.path(workDir,"multi_initializeMainAlgo.R"))
source(file.path(workDir,"multi_estep.R"))
source(file.path(workDir,"multi_mstep.R"))
source(file.path(workDir,"multi_main.R"))
source(file.path(workDir,"multi_map.R"))
source(file.path(workDir,"compute_LL_multi.R"))
```

#### Formatting data 
The function saemixData() requires a mandatory argument which is the name of the dataset. The dataset has to be formatted in order to obtain an id column corresponding to the patient id, a time column corresponding to the sampling times, an observation column corresponding to the response observations and a ytype column corresponding to the distinct response types. Optional columns can be added if user the wants to model covariates for example. See saemix documentation for more details. 
In this example, prothrombin measurements correspond to response 1 (ytype = 1) and survival measurements correspond to response 2 (ytype = 2). For the survival reponse (ytype = 2), observation is 1 in case of event (death) and 0 otherwise. 

```{r Data}
d1 = prothro[,c(1,2,3)]
d1$ytype=1
colnames(d1)[2] = "obs"
d2 = prothros[,c(1,3,2)]
d2$ytype = 2
colnames(d2)[2] = "obs"
colnames(d2)[3] = "time"
data_joint = rbind(d1,d2)
# To see the data for patient 1 
data_joint[data_joint$id==1,]
```

The user is encouraged to specify optional arguments of the saemixData() function: the id variable (name.group argument), the predictor variables (name.predictors argument) with at least the sampling times, the observation variable (name.reponse argument) and the response type variable (name.ytype argument).

```{r saemixData}
saemix.data<-saemixData(name.data=data_joint, name.group=c("id"), 
                        name.predictors=c("time","obs"),
                        name.response="obs",name.ytype = "ytype")
```


#### Joint model with a linear mixed-effects model and a survival model with constant baseline hazard


The saemixModel() function requires two mandatory arguments. The first one is a R function describing the joint model involving the structural model for longitudinal observations and the likelihood contribution for the survival observations. The second one is a matrix with a number of columns equal to the number of parameters, and one (when no covariates) or several row (when covariates enter the model) giving the initial estimates of fixed-effects. 
The user is encouraged to specify optional arguments of the saemixModel() function: the response type with the modeltype argument ("structural" for longitudinal observations and "likelihood" for survival ones), the distribution of each parameter with the transform.par argument (0 = normal, 1 = log-normal, 2 = probit and 3 = logit), the fixed or estimated parameters with the fixed.estim argument (0 = to be fixed to the initial estimate, 1 = to be estimated), if random effects are added with the covariance.model argument (square matrix of size equal to the number of parameters giving the variance-covariance matrix of the model), the initialization of random effect variances with the omega.init argument (square matrix of size equal to the number of parameters giving the initialization of the variance-covariance matrix of the model), the error model with the error.model argument (valid types are "constant", "proportional", "combined"). Futher arguments can be considered and found in the package description (see documentation). 

In the following we start with a simple case: a joint model with a linear mixed-effects and a survival model involving constant baseline risk. The joint model writes: 

\begin{align}
     y_{ij} = & m_l(t_{ij},\psi_{i}) + g[m_l(t_{ij},\psi_{i}), \sigma]\epsilon_{ij} \notag \\
      = & \psi_{i0} + \psi_{i1}\times t_{ij} + \sigma\epsilon_{ij} \notag \\
     h_i(t, \psi_i) = & h_0 \, \exp(\alpha \, m(t,\psi_i))
\end{align}

We then define the model to be entered in the function. This function must have 3 arguments named psi (assumed to be a matrix with the number of columns equal to the number of parameters in the model (excluding error parameters), so here 4 for $\mu_0$, $\mu_1$, $h_0$ and $\alpha$), id (assumed to be a vector of indices matching observation number with subject index) and xidep (assumed to be a matrix with as many columns as predictors + 1 for the type of response, so here 3 for time, observations and response type). The three arguments passed to the function will be generated automatically from the model and data object within the saemix code. The
function must return a vector composed of predictions for longitudinal responses and likelihood contributions for the survival responses. The length of the vector is equal to the number of rows in the predictor xidep. 

```{r model definition}

JMmodel<-function(psi,id,xidep) {
  ytype<-xidep$ytype  # type of response (1: continuous, 2: event)
  b0 <- psi[id,1] 
  b1 <- psi[id,2] 
  h0 <- psi[id,3]
  alpha <- psi[id,4] 
  
  ypred <- b0+b1*xidep$time # predictions for the longitudinal part
  
  T<-xidep$time[ytype==2] # vector of times (survival response)
  Nj <- length(T)
  ev = xidep$obs[ytype==2] # vector of observations (survival response)
  cens<-which(ev==0)       # with censored ones
  ind <- which(ev==1)      # and event ones 
  
  # Creating vectors of the same length of T to compute likelihood of the survival part 
  #(so removing duplicates)
  b0b = b0[ytype==2] # to have vectors of the same length as T 
  b1b = b1[ytype==2]
  h0b = h0[ytype==2]
  alphab = alpha[ytype==2]
  
  haz <- h0b*exp(alphab*(b0b+b1b*T)) # instantaneous hazard
  # cumulative hazard (explicit expression in that case)
  H <- (h0b/(alphab*b1b))*exp((b0b+b1b*T)*alphab)-(h0b/(alphab*b1b))*exp(alphab*b0b) 
  
  logpdf <- rep(0,Nj)
  logpdf[cens] <- -H[cens] # likelihood contributions for censored observations
  logpdf[ind] <- -H[ind] + log(haz[ind]) # likelihood contributions for event observations 
  
  ypred[ytype==2] = logpdf
  return(ypred)
}

```

```{r saemixModel}

#### initializing parameters

param<-c(73,1.25,0.6,0.0001) 
omega.sim<-c(18, 3, 0.05, 0.01)
sigma.sim <- 17

### saemix Model 

saemix.model<-saemixModel(model=JMmodel,description="JM LMEM-TTE constant baseline hazard", 
                      modeltype=c("structural","likelihood"),
                      psi0=matrix(param,ncol=4,byrow=TRUE,
                                  dimnames=list(NULL, c("b0","b1","h0","alpha"))),
                      transform.par=c(0,0,1,0), covariance.model=diag(c(1,1,0,0)),
                      fixed.estim = c(1,1,1,1),error.model = "constant",
                      omega.init = diag(omega.sim))


```

In the following we specify some algorithm settings. The option fim = T is specified to obtain standard errors of parameter estimates. ll.is is specified to obtain the loglikelihood at the MLE, the AIC and BIC. Graphs are not currently adapted so please specify save.graphs = F. We run the algorithm using the saemix.multi() function. 

```{r saemix options}
saemix.options<-saemixControl(seed=12345, map=T, fim=T, ll.is=TRUE, save.graphs = F) 
# please, specify save.graphs=F (currently not extended)
yfit <- saemix.multi(saemix.model, saemix.data, saemix.options)
```

We obtain the summary of the fit with the parameter estimates, and likelihood value at MLE (with AIC and BIC).
The standard error estimates are obtained using the following script. 

```{r saemix results}
yfit@results@fim  # variance covariance matrix (inverse of the FIM)
sqrt(diag(yfit@results@fim)) # standard errors of parameters estimates
# Formatting results in a data frame
d = data.frame(par=c(yfit@results@name.fixed,yfit@results@name.random,"sigma_a1"),
               est = c(yfit@results@fixed.effects,diag(yfit@results@omega)[1:2],
                       yfit@results@respar[c(1)]), 
               se = sqrt(diag(yfit@results@fim)))
print(d)
```

## COVID-19 example (multivariate joint model with competing risks)

We consider the joint model defined in the article involving three biomarkers and two competing risks: 

\begin{align}
    y_{ij1} &= m_1(t_{ij1},\psi_{i1}) + \sigma_1 \, m_1(t_{ij1},\psi_{i1})\epsilon_{ij1}\\
    y_{ij2} &= m_{2}(t_{ij2},\psi_{i2}) + \sigma_{2} \epsilon_{ij2} \notag \\
    y_{ij3} &= m_{3}(t_{ij3},\psi_{i3}) + \sigma_{3} \epsilon_{ij3} \notag \\
    \notag \\
    h_{1}(t,\psi_i,Score_{i};\theta) &=  h_0 \exp \left[\sum_{k=1}^{3} \Big(\alpha_{k} \times \big(m_k(t,\psi_{ik})-med_k\big)\Big) + \beta \cdot Score_{i} \right] \notag\\
    h_{2}(t,\psi_i,Score_{i};\theta) & =  \frac{1}{b} \frac{\left( 1-F_1(\infty)\right) \exp(-t/b)}{1-\left( 1-F_1(\infty)\right) \left(1-\exp(-t/b)\right)} \notag
\end{align}
with:
\begin{align}
    m_1(t_{ij1},\psi_{i1}) & = \psi_{i01} + \psi_{ia1} \times \left[ \exp\left(\psi_{i11}\, t_{ij1}\right)-\exp\left(\psi_{i21}\, t_{ij1}\right) \right]  \\
    m_{2}(t_{ij2},\psi_{i2}) & = \psi_{i02} + \psi_{i12}\times t_{ij2} \notag \\
    m_{3}(t_{ij3},\psi_{i3}) & = \psi_{i03} + \psi_{i13}\times t_{ij3} \notag
\end{align}


#### Creating the saemix data object 
```{r saemixData 2}
data_joint = read.table("W:/saemix/dev/riscov/dataJM.txt", header = T)
dataJM<-saemixData(name.data=data_joint, name.group=c("id"), name.predictors=c("time","obs"),
                   name.response="obs", name.ytype = "ytype", name.covariates = c("score4C"))
```

#### Creating the saemix model object 
The model can be implemented as follows. 
```{r model definition 2}

JMmodel<-function(psi,id,xidep) {
    ytype<-xidep$ytype  
    N = unique(id)
   
    b01 <- psi[id,1]
    b11 <- psi[id,2]
    b21 <- psi[id,3]
    a1 <- psi[id,4]
    b02 <- psi[id,5]
    b12 <- psi[id,6]
    b03 <- psi[id,7]
    b13 <- psi[id,8]
    h1 <- psi[id,9] 
    alpha1 <- psi[id,10]
    alpha2 <- psi[id,11]
    alpha3 <- psi[id,12]
    b <- psi[id,13]
    cov <- psi[id,14]
  
    T2 = xidep[ytype==2,1]  # vector of times for biom 2
    T3 = xidep[ytype==3,1]  # vector of times for biom 3
    T<-xidep[ytype==4,1]    # vector of times partie survie
    ev = xidep$obs[ytype==4]
    Nj <- length(T)
    cens<-which(ev==0)  # index of censored observations
    ind1 <- which(ev==1) # index of event 1
    ind2 <- which(ev==2) # index of event 2

    schem = sapply(N, function(i) length(which(id <= i))) 
    # schem is a vector giving the number of measurements for each patients
    # defined for having unique individual parameters (see what follows)
  
    b01b <- b01[schem]
    b11b <- b11[schem]
    b21b = b21[schem]
    a1b = a1[schem]
    b02b <- b02[schem]
    b12b <- b12[schem]
    b03b = b03[schem]
    b13b = b13[schem]
    h1b <- h1[schem]
    alpha1b = alpha1[schem]
    alpha2b = alpha2[schem]
    alpha3b = alpha3[schem]
    bb = b[schem]
    covb = cov[schem]
    
    f=function(x) seq(0,x,length.out=100)
    tab = mapply(f,T)
    tab = t(tab)
    pas = tab[,2]-tab[,1]
    # used for numerical integration because no explicit form of the cumulative hazard
  
    f2=function(x) seq(0,x,length.out=1000)
    tab2 = replicate(Nj,f2(1000))
    tab2 = t(tab2)
    pas2 = tab2[,2]-tab2[,1]
    # used to compute the value F1(inf) (used in the competing event hazard)
    # proxy: F1(inf) approximated by F1(1000)
  
    haz1 = h1b*exp(alpha1b*(b01b+a1b*(exp(b11b*tab)-exp(b21b*tab))-5.78)
                   +alpha2b*(b02b+b12b*tab-7.42)
                   +alpha3b*(b03b+b13b*tab-2.89)+covb)
    H1 = apply(haz1,1,sum)*pas # cumulative hazard from 0 to T
    hazt1 = haz1[,100] # cumulative hazard for event 1 at time T
  
    haz1b = h1b*exp(alpha1b*(b01b+a1b*(exp(b11b*tab2)-exp(b21b*tab2))-5.78)
                    +alpha2b*(b02b+b12b*tab2-7.42)
                    +alpha3b*(b03b+b13b*tab2-2.89)+covb)
    H1b = apply(haz1b,1,sum)*pas2 
    F1 = 1-exp(-H1b) # F1(1000)
   
    hazt2 = 1/bb*(1-F1)*exp(-T/bb)/(1-(1-F1)*(1-exp(-T/bb)))
    H2 = -log(1-(1-F1)*(1-exp(-T/bb))) # cumulative hazard for event 2 at time T
  
    logpdf <- rep(0,Nj)
    logpdf[cens] <- log(exp(-H1[cens])+exp(-H2[cens])-1) 
    # likelihood contributions for censored observations
    logpdf[ind1] <- -H1[ind1] + log(hazt1[ind1]) 
    # likelihood contributions for event 1 observations
    logpdf[ind2] <- -H2[ind2] + log(hazt2[ind2]) 
    # likelihood contributions for event 2 observations
  
    ypred = rep(NA,length(xidep[,1]))
  
    ypred[ytype==1] = b01[ytype==1]+a1[ytype==1]*(exp(b11[ytype==1]*xidep[ytype==1,1])
                                                  -exp(b21[ytype==1]*xidep[ytype==1,1]))
    ypred[ytype==2] = b02[ytype==2]+b12[ytype==2]*xidep[ytype==2,1]
    ypred[ytype==3] = b03[ytype==3]+b13[ytype==3]*xidep[ytype==3,1]
  
    ypred[ytype==4] = logpdf
  
    return(ypred)
}

# Initial guesses for parameter estimates 
param_longi<-c(4.5,-0.14,-0.16,6.1,7.4,0.0008,4.2,-0.15) 
omega_longi <- c(4,0.01,0.003,0.7,0.001,0.00001,0.8,0.01)
param_surv<-c(0.0009,0.25,-9,0.76,16,0)
omega_surv <- c(0.01,0.01,0.01,0.01,0.01,0.01) 
param_cov <-  c(0,0,0,0,0,0,0,0,0,0,0,0,0,0.3)
param <- matrix(data = c(param_longi, param_surv, param_cov), ncol = 14, byrow = T, 
                dimnames=list(NULL, c( "b01", "b11", "b21", "a1","b02", "b12", "b03", "b13", 
                                       "h1", "alpha1", "alpha2", "alpha3", "b", "cov")))
omega <- c(omega_longi, omega_surv) 
  
jointTTE<-saemixModel(model=JMmodel,description="JM lin+competing risks",
                      modeltype=c("structural","structural","structural","likelihood"),
                      psi0=param,transform.par=c(0,0,0,1,0,0,0,0,1,0,0,0,1,0), 
                      covariance.model=diag(c(1,1,1,1,1,1,1,1,0,0,0,0,0,0)),
                      omega.init = diag(omega), fixed.estim = c(1,1,1,1,1,1,1,1,1,1,1,1,1,0),
                      covariate.model = c(0,0,0,0,0,0,0,0,0,0,0,0,0,1),
                      error.model = c("proportional","constant","constant","likelihood"))


```
