---
title: "Package compilation for saemix 3.5 and basic run"
author: "Emmanuelle Comets"
date: "25/02/2026"
output:
  pdf_document: null
  latex_engine: xelatex
  word_document: default
html_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(devtools)
library(roxygen2)
library(tools) # check reverse dependencies
saemixDir<-"/home/eco/work/saemix/saemixextension"
workDir<-"/home/eco/work/saemix/versions/saemix3.5"
knitr::opts_chunk$set(root.dir = workDir)

# Options
testExamples <- TRUE
checkReverseDependencies <- TRUE
installPackage <- FALSE

```

## Copy files

```{r copyFileDev}
cmd<-paste("rm -r ",file.path(workDir,"*"),sep="")
system(cmd)

dir.create(workDir)
dir.create(file.path(workDir,"saemix"))

cmd<-paste("cp -rp ",file.path(saemixDir,"documentation","docsaem.pdf")," ",file.path(saemixDir,"inst","userguide_saemix3.pdf"),sep="")
system(cmd)

cmd<-paste("cp -rp ",file.path(saemixDir,"R")," ", file.path(workDir,"saemix","/"),sep="")
system(cmd)
cmd<-paste("cp -rp ",file.path(saemixDir,"data")," ", file.path(workDir,"saemix","/"),sep="")
system(cmd)
cmd<-paste("cp -rp ",file.path(saemixDir,"inst")," ", file.path(workDir,"saemix","/"),sep="")
system(cmd)
for(ifile in c("CHANGES","DESCRIPTION")) {
  cmd<-paste("cp ",file.path(saemixDir,ifile)," ", file.path(workDir,"saemix"),sep="")
  system(cmd)
}
cmd<-paste("cp ",file.path(saemixDir,"inst","CITATION")," ", file.path(workDir,"saemix","inst"),sep="")
system(cmd)

```
## Submission message

This is an udpate to the saemix package.

Please ignore the warning concerning the following "possibly mis-spelled words in DESCRIPTION:"
  IAME (17:930)
  al (17:663)
  et (17:660)
  github (17:954)
  saemix (17:965)

The examples in this package have been commented out for the CRAN submission as they take too long to run, but they have been checked before submission.

Reverse dependencies have also been checked.

## Compilation

- building on win-ftp
  - to specify when submitting
Possibly mis-spelled words in DESCRIPTION:
  IAME (17:930)
  al (17:663)
  et (17:660)
  github (17:954)
  saemix (17:965)

  - removed:
Found the following (possibly) invalid URLs:
  URL: http://group.monolix.org/
    From: DESCRIPTION
    Status: 403
    Message: Forbidden

  - to change: no commas between keywords in R (maybe for vignettes)
  - previous version, solved now:
````markdown
`r ''````{r}
Found the following \keyword or \concept entries
which likely give several index terms:
  File 'backward.procedure.Rd':
    \keyword{backward,}
    \keyword{selection,}
... (others)
```
````
- new NOTE (October 2023)

* checking HTML version of manual ... NOTE
Skipping checking HTML validation: no command 'tidy' found

- compilation to pdf (as comments)

<!-- Messages d'avis : -->
<!-- 1: Dans do_once((if (is_R_CMD_check()) stop else warning)("The function xfun::isFALSE() will be deprecated in the future. Please ",  : -->
<!--   The function xfun::isFALSE() will be deprecated in the future. Please consider using base::isFALSE(x) or identical(x, FALSE) instead. -->
<!-- 2: Dans in_dir(input_dir(), expr) : -->
<!--   You changed the working directory to /home/eco/work/saemix/versions/saemix3.3/saemix (probably via setwd()). It will be restored to /home/eco/work/saemix/saemixextension/compilation. See the Note section in ?knitr::knit -->
<!-- 3: Dans in_dir(input_dir(), expr) : -->
<!--   You changed the working directory to /home/eco/work/saemix/versions/saemix3.3 (probably via setwd()). It will be restored to /home/eco/work/saemix/saemixextension/compilation. See the Note section in ?knitr::knit -->
<!-- 4: Dans xfun::read_utf8(log) : -->
<!--   The file compileSaemix3.3.log is not encoded in UTF-8. These lines contain invalid UTF-8 characters: 526, 533 -->
<!-- 5: Dans grep("^(LaTeX|Package [[:alnum:]]+) Warning:", x) : -->
<!--   unable to translate '/10 a <e9>t<e9> trouv<e9>e dans les packages et les biblioth<e8>ques :[] ' to a wide string -->
<!-- 6: Dans grep("^(LaTeX|Package [[:alnum:]]+) Warning:", x) : -->
<!--   input string 526 is invalid -->
<!-- 7: Dans grep("^(LaTeX|Package [[:alnum:]]+) Warning:", x) : -->
<!--   unable to translate ' <e9>t<e9> trouv<e9>e dans les packages et les biblioth<e8>ques :[] ' to a wide string -->
<!-- 8: Dans grep("^(LaTeX|Package [[:alnum:]]+) Warning:", x) : -->
<!--   input string 533 is invalid -->


```{r document, echo=FALSE, results=FALSE, message=FALSE, warning=FALSE}
# Reduce size of viral load simulation data (to 500 simulations) for CRAN
knitr::opts_chunk$set(root.dir = file.path(workDir,"saemix"))
setwd(file.path(workDir,"saemix"))

devtools::document(file.path(workDir,"saemix"))
#source(file.path(workDir,"saemix","R","aaa_generics.R"))
roxygenise(package.dir=file.path(workDir,"saemix"))
```

Current dependencies, versions downloaded on 29/07/25:
- varTestnlme
- nlive
- mkin

```{r compilation}
setwd(workDir)
system("R CMD build saemix")

# Test examples
if(testExamples) 
  system("R CMD check --as-cran --run-donttest saemix_3.5.tar.gz") else 
    system("R CMD check --as-cran saemix_3.5.tar.gz")

# Reverse dependencies
cmd<-paste("cp ",file.path("/home/eco/work/saemix/versions/reverseDependencies/varTestnlme_1.3.5.tar.gz")," ", file.path(workDir),sep="")
system(cmd)
cmd<-paste("cp ",file.path("/home/eco/work/saemix/versions/reverseDependencies/nlive_0.8.0.tar.gz")," ", file.path(workDir),sep="")
system(cmd)
cmd<-paste("cp ",file.path("/home/eco/work/saemix/versions/reverseDependencies/mkin_1.2.10.tar.gz")," ", file.path(workDir),sep="")
system(cmd)

if(checkReverseDependencies) {
  # Check which packages depend on saemix
  package_dependencies(packages="saemix", reverse=TRUE)
  # TODO download these packages into workDir
}

if(installPackage) {  
  # install saemix current version and check packages
  install.packages(pkgs=file.path(workDir,"saemix_3.5.tar.gz"),repos=NULL)
  #result <- check_packages_in_dir(workDir, revdep = list() )
  result <- check_packages_in_dir(workDir, revdep = list("varTestnlme") )
  summary(result)
}
```

#### Examples on CRAN

Examples with CPU (user + system) or elapsed time > 5s
                    user system elapsed
saemix.bootstrap 250.647  0.944 251.698
rapi.saemix       32.251  0.159  32.416
cow.saemix        18.148  0.634  18.817
PD1.saemix        13.211  0.147  13.369
toenail.saemix    10.560  0.003  10.564
compare.saemix     7.266  0.015   7.353
yield.saemix       5.602  0.011   5.612
theo.saemix        5.585  0.007   5.593

### Warnings

* checking for future file timestamps ... NOTE
unable to verify current time

* checking HTML version of manual ... NOTE
Skipping checking HTML validation: no command 'tidy' found

* checking examples ... [386s/387s] NOTE
Examples with CPU (user + system) or elapsed time > 5s

## Check

### Install package in development mode

```{r devModeInstall}
dev_mode() # development mode
install.packages(pkgs=file.path(workDir,"saemix_3.5.tar.gz"),repos=NULL)
library(saemix)
library(testthat)
```

### Running theopp example

For some weird reason, this does not work as of 3.5 (man pages are not found in dev_mode, or in this session after the code above has been run, but they are in the package and can be used in a new session if the library is installed...)

```{r helpTheo}
?theo.saemix
?saemix
```

Run on the theophylline example (works, same results as documentation (corrected error in covariate model, was on V instead of CL as in the doc)

Note: p-value of .38 instead of .19 in the doc for beta, not sure why as estimates didn't change

```{r runTheo, echo=FALSE}
data(theo.saemix)

saemix.data<-saemixData(name.data=theo.saemix,header=TRUE,sep=" ",na=NA,
   name.group=c("Id"),name.predictors=c("Dose","Time"),
   name.response=c("Concentration"),name.covariates=c("Weight","Sex"),
   units=list(x="hr",y="mg/L", covariates=c("kg","-")), name.X="Time", verbose=FALSE)
print(saemix.data)
plot(saemix.data)

model1cpt<-function(psi,id,xidep) { 
	  dose<-xidep[,1]
	  tim<-xidep[,2]  
	  ka<-psi[id,1]
	  V<-psi[id,2]
	  CL<-psi[id,3]
	  k<-CL/V
	  ypred<-dose*ka/(V*(ka-k))*(exp(-k*tim)-exp(-ka*tim))
	  return(ypred)
}

saemix.model<-saemixModel(model=model1cpt,
  description="One-compartment model with first-order absorption", 
  psi0=matrix(c(1.,20,0.5,0.1,0,-0.01),ncol=3, byrow=TRUE,
  dimnames=list(NULL, c("ka","V","CL"))),transform.par=c(1,1,1),
  covariate.model=matrix(c(0,0,1,0,0,0),ncol=3,byrow=TRUE),fixed.estim=c(1,1,1),
  covariance.model=matrix(c(1,0,0,0,1,0,0,0,1),ncol=3,byrow=TRUE),
  omega.init=matrix(c(1,0,0,0,1,0,0,0,1),ncol=3,byrow=TRUE),error.model="constant", verbose=FALSE)
print(saemix.model)

saemix.fit<-saemix(model=saemix.model,
                   data=saemix.data,
                   control=list(seed=632545,directory="newtheo", save=FALSE,save.graphs=FALSE, warnings=FALSE))

# saemix.fit2<-saemix(model=saemix.model, data=saemix.data, control=list(seed=632545,directory="newtheo", save=FALSE,save.graphs=FALSE, warnings=FALSE, nb.chains=1))

# Prints a summary of the results
print(saemix.fit)

# New function, same seed, same results
sim2<-simulate.SaemixObject(saemix.fit, seed=12345)
sim2@sim.data

# Outputs the estimates of individual parameters
psi(saemix.fit)

# legacy
if(FALSE){
  sim1<-simul.saemix(saemix.fit, seed=12345)
  # no simulated data
  saemix.fit@sim.data
  # simulated data
  sim1@sim.data
  identical(sim1@sim.data, sim2@sim.data)
}
```

Plot results:

```{r plotTheo, echo=FALSE}
# Shows some diagnostic plots to evaluate the fit
plot(saemix.fit)
plot(saemix.fit, plot.type="vpc")
plot(saemix.fit, plot.type="individual")

# Routes to npdeSaemix
plot(saemix.fit, plot.type="npde")
ynpde<-npdeSaemix(saemix.fit)
# ecdf by weight
plot(ynpde, plot.type="ecdf", which.cov="Weight", covsplit=T)
# scatterplot by sex - not working, Sex not in ynde...
plot(ynpde, plot.type="x.scatter", which.cov="Sex", covsplit=T)

```

Bootstrap

```{r theoBoot}

```


```{r deactivateDevMode, echo=FALSE, hide=T}
dev_mode()
```

