---
title: "Package compilation for saemix 3.3 and basic run"
author: "Emmanuelle Comets"
date: "15/10/2023"
output:
  pdf_document: null
  latex_engine: xelatex
  word_document: default
html_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(devtools)
library(roxygen2)
library(tools) # check reverse dependencies
saemixDir<-"/home/eco/work/saemix/saemixextension"
workDir<-"/home/eco/work/saemix/versions/saemix3.3"
knitr::opts_chunk$set(root.dir = workDir)

# Options
testExamples <- TRUE
checkReverseDependencies <- TRUE
installPackage <- FALSE

```

## Copy files

```{r copyFileDev}
cmd<-paste("rm -r ",file.path(workDir,"*"),sep="")
system(cmd)

dir.create(workDir)
dir.create(file.path(workDir,"saemix"))

cmd<-paste("cp -rp ",file.path(saemixDir,"documentation","docsaem.pdf")," ",file.path(saemixDir,"inst","userguide_saemix3.pdf"),sep="")
system(cmd)

cmd<-paste("cp -rp ",file.path(saemixDir,"R")," ", file.path(workDir,"saemix","/"),sep="")
system(cmd)
cmd<-paste("cp -rp ",file.path(saemixDir,"data")," ", file.path(workDir,"saemix","/"),sep="")
system(cmd)
cmd<-paste("cp -rp ",file.path(saemixDir,"inst")," ", file.path(workDir,"saemix","/"),sep="")
system(cmd)
for(ifile in c("CHANGES","DESCRIPTION")) {
  cmd<-paste("cp ",file.path(saemixDir,ifile)," ", file.path(workDir,"saemix"),sep="")
  system(cmd)
}
cmd<-paste("cp ",file.path(saemixDir,"inst","CITATION")," ", file.path(workDir,"saemix","inst"),sep="")
system(cmd)

```

## Compilation

- building on win-ftp
  - to specify when submitting
Possibly mis-spelled words in DESCRIPTION:
  IAME (17:930)
  al (17:663)
  et (17:660)
  github (17:954)
  saemix (17:965)

  - removed:
Found the following (possibly) invalid URLs:
  URL: http://group.monolix.org/
    From: DESCRIPTION
    Status: 403
    Message: Forbidden

  - to change: no commas between keywords in R (maybe for vignettes)
  - previous version, solved now:
````markdown
`r ''````{r}
Found the following \keyword or \concept entries
which likely give several index terms:
  File 'backward.procedure.Rd':
    \keyword{backward,}
    \keyword{selection,}
... (others)
```
````

- Examples with CPU or elapsed time > 5s
                 user system elapsed
rapi.saemix    47.006  0.096  47.115
cow.saemix     25.185  0.120  25.311
PD1.saemix     18.925  0.048  19.108
toenail.saemix 16.656  0.020  16.676
compare.saemix  9.152  0.020   9.171
theo.saemix     6.481  0.008   6.490
yield.saemix    5.419  0.016   5.435

- new NOTE (October 2023)

* checking HTML version of manual ... NOTE
Skipping checking HTML validation: no command 'tidy' found

- compilation to pdf (as comments)

<!-- Messages d'avis : -->
<!-- 1: Dans do_once((if (is_R_CMD_check()) stop else warning)("The function xfun::isFALSE() will be deprecated in the future. Please ",  : -->
<!--   The function xfun::isFALSE() will be deprecated in the future. Please consider using base::isFALSE(x) or identical(x, FALSE) instead. -->
<!-- 2: Dans in_dir(input_dir(), expr) : -->
<!--   You changed the working directory to /home/eco/work/saemix/versions/saemix3.3/saemix (probably via setwd()). It will be restored to /home/eco/work/saemix/saemixextension/compilation. See the Note section in ?knitr::knit -->
<!-- 3: Dans in_dir(input_dir(), expr) : -->
<!--   You changed the working directory to /home/eco/work/saemix/versions/saemix3.3 (probably via setwd()). It will be restored to /home/eco/work/saemix/saemixextension/compilation. See the Note section in ?knitr::knit -->
<!-- 4: Dans xfun::read_utf8(log) : -->
<!--   The file compileSaemix3.3.log is not encoded in UTF-8. These lines contain invalid UTF-8 characters: 526, 533 -->
<!-- 5: Dans grep("^(LaTeX|Package [[:alnum:]]+) Warning:", x) : -->
<!--   unable to translate '/10 a <e9>t<e9> trouv<e9>e dans les packages et les biblioth<e8>ques :[] ' to a wide string -->
<!-- 6: Dans grep("^(LaTeX|Package [[:alnum:]]+) Warning:", x) : -->
<!--   input string 526 is invalid -->
<!-- 7: Dans grep("^(LaTeX|Package [[:alnum:]]+) Warning:", x) : -->
<!--   unable to translate ' <e9>t<e9> trouv<e9>e dans les packages et les biblioth<e8>ques :[] ' to a wide string -->
<!-- 8: Dans grep("^(LaTeX|Package [[:alnum:]]+) Warning:", x) : -->
<!--   input string 533 is invalid -->


```{r document, echo=FALSE, results=FALSE, message=FALSE, warning=FALSE}
# Reduce size of viral load simulation data (to 500 simulations) for CRAN
knitr::opts_chunk$set(root.dir = file.path(workDir,"saemix"))
setwd(file.path(workDir,"saemix"))

devtools::document(file.path(workDir,"saemix"))
roxygenise()
```

Current dependencies, versions downloaded on 18/10/2023:
- varTestnlme
- nlive
- mkin

```{r compilation}
setwd(workDir)
system("R CMD build saemix")

# Test examples
if(testExamples) 
  system("R CMD check --as-cran --run-donttest saemix_3.3.tar.gz") else 
    system("R CMD check --as-cran saemix_3.3.tar.gz")

# Reverse dependencies
cmd<-paste("cp ",file.path("/home/eco/work/saemix/versions/reverseDependencies/varTestnlme_1.3.5.tar.gz")," ", file.path(workDir),sep="")
system(cmd)
cmd<-paste("cp ",file.path("/home/eco/work/saemix/versions/reverseDependencies/nlive_0.1.0.tar.gz")," ", file.path(workDir),sep="")
system(cmd)
cmd<-paste("cp ",file.path("/home/eco/work/saemix/versions/reverseDependencies/mkin_1.2.6.tar.gz")," ", file.path(workDir),sep="")
system(cmd)

if(checkReverseDependencies) {
  # Check which packages depend on saemix
  package_dependencies(packages="saemix", reverse=TRUE)
  # TODO download these packages into workDir
}

if(installPackage) {  
  # install saemix current version and check packages
  install.packages(pkgs=file.path(workDir,"saemix_3.3.tar.gz"),repos=NULL)
  #result <- check_packages_in_dir(workDir, revdep = list() )
  result <- check_packages_in_dir(workDir, revdep = list("varTestnlme") )
  summary(result)
}
```


#### Examples on CRAN

Examples with CPU or elapsed time > 5s
                 user system elapsed
rapi.saemix    32.510  0.092  32.602
cow.saemix     17.020  0.072  17.093
toenail.saemix 13.531  0.000  13.532
PD1.saemix     12.789  0.035  12.910
compare.saemix  6.331  0.016   6.347
theo.saemix     5.171  0.000   5.172

### Warnings

## Check

### Install package in development mode
  
```{r devModeInstall}
dev_mode() # development mode
install.packages(pkgs=file.path(workDir,"saemix_3.3.tar.gz"),repos=NULL)
library(saemix)
library(testthat)
```

### Running theopp example

```{r helpTheo}
?theo.saemix
?saemix
```

Run on the theophylline example

```{r runTheo, echo=FALSE}
data(theo.saemix)

saemix.data<-saemixData(name.data=theo.saemix,header=TRUE,sep=" ",na=NA,
   name.group=c("Id"),name.predictors=c("Dose","Time"),
   name.response=c("Concentration"),name.covariates=c("Weight","Sex"),
   units=list(x="hr",y="mg/L", covariates=c("kg","-")), name.X="Time", verbose=FALSE)
print(saemix.data)
plot(saemix.data)

model1cpt<-function(psi,id,xidep) { 
	  dose<-xidep[,1]
	  tim<-xidep[,2]  
	  ka<-psi[id,1]
	  V<-psi[id,2]
	  CL<-psi[id,3]
	  k<-CL/V
	  ypred<-dose*ka/(V*(ka-k))*(exp(-k*tim)-exp(-ka*tim))
	  return(ypred)
}

saemix.model<-saemixModel(model=model1cpt,
  description="One-compartment model with first-order absorption", 
  psi0=matrix(c(1.,20,0.5,0.1,0,-0.01),ncol=3, byrow=TRUE,
  dimnames=list(NULL, c("ka","V","CL"))),transform.par=c(1,1,1),
  covariate.model=matrix(c(0,1,0,0,0,0),ncol=3,byrow=TRUE),fixed.estim=c(1,1,1),
  covariance.model=matrix(c(1,0,0,0,1,0,0,0,1),ncol=3,byrow=TRUE),
  omega.init=matrix(c(1,0,0,0,1,0,0,0,1),ncol=3,byrow=TRUE),error.model="constant", verbose=FALSE)
print(saemix.model)

saemix.fit<-saemix(model=saemix.model,
                   data=saemix.data,
                   control=list(seed=632545,directory="newtheo", save=FALSE,save.graphs=FALSE, warnings=FALSE))

# saemix.fit2<-saemix(model=saemix.model, data=saemix.data, control=list(seed=632545,directory="newtheo", save=FALSE,save.graphs=FALSE, warnings=FALSE, nb.chains=1))

# Prints a summary of the results
print(saemix.fit)

# New function, same seed, same results
sim2<-simulate.SaemixObject(saemix.fit, seed=12345)
sim2@sim.data

# Outputs the estimates of individual parameters
psi(saemix.fit)

# legacy
if(FALSE){
  sim1<-simul.saemix(saemix.fit, seed=12345)
  # no simulated data
  saemix.fit@sim.data
  # simulated data
  sim1@sim.data
  identical(sim1@sim.data, sim2@sim.data)
}
```

Plot results:

```{r plotTheo, echo=FALSE}
# Shows some diagnostic plots to evaluate the fit
plot(saemix.fit)
plot(saemix.fit, plot.type="vpc")
plot(saemix.fit, plot.type="individual")

# Routes to npdeSaemix
plot(saemix.fit, plot.type="npde")
ynpde<-npdeSaemix(saemix.fit)
# ecdf by weight
plot(ynpde, plot.type="ecdf", which.cov="Weight", covsplit=T)
# scatterplot by sex - not working, Sex not in ynde...
plot(ynpde, plot.type="x.scatter", which.cov="Sex", covsplit=T)

```

Bootstrap

```{r theoBoot}

```


```{r deactivateDevMode, echo=FALSE, hide=T}
dev_mode()
```

