---
title: "Package compilation for saemix 3.0 and basic run"
author: "Emmanuelle Comets"
date: "18/10/2021"
output:
  pdf_document:
  latex_engine: xelatex
html_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(devtools)
library(roxygen2)
saemixDir<-"/home/eco/work/saemix/saemixextension"
workDir<-"/home/eco/work/saemix/versions/saemix3.0"
knitr::opts_chunk$set(root.dir = workDir)
```

## Copy files

```{r copyFileDev}
cmd<-paste("rm -r ",file.path(workDir,"*"),sep="")
system(cmd)

dir.create(workDir)
dir.create(file.path(workDir,"saemix"))

cmd<-paste("cp -rp ",file.path(saemixDir,"documentation","docsaem.pdf")," ",file.path(saemixDir,"inst","userguide_saemix3.pdf"),sep="")
system(cmd)

cmd<-paste("cp -rp ",file.path(saemixDir,"R")," ", file.path(workDir,"saemix","/"),sep="")
system(cmd)
cmd<-paste("cp -rp ",file.path(saemixDir,"data")," ", file.path(workDir,"saemix","/"),sep="")
system(cmd)
cmd<-paste("cp -rp ",file.path(saemixDir,"inst")," ", file.path(workDir,"saemix","/"),sep="")
system(cmd)
for(ifile in c("CHANGES","DESCRIPTION")) {
  cmd<-paste("cp ",file.path(saemixDir,ifile)," ", file.path(workDir,"saemix"),sep="")
  system(cmd)
}
cmd<-paste("cp ",file.path(saemixDir,"inst","CITATION")," ", file.path(workDir,"saemix","inst"),sep="")
system(cmd)


```

## Compilation

- building on win-ftp
  - to specify when submitting
Possibly mis-spelled words in DESCRIPTION:
  IAME (17:930)
  al (17:663)
  et (17:660)
  github (17:954)
  saemix (17:965)

  - removed:
Found the following (possibly) invalid URLs:
  URL: http://group.monolix.org/
    From: DESCRIPTION
    Status: 403
    Message: Forbidden

  - to change: no commas between keywords in R (maybe for vignettes)
Found the following \keyword or \concept entries
which likely give several index terms:
  File 'backward.procedure.Rd':
    \keyword{backward,}
    \keyword{selection,}
... (others)

```{r compilation}
# Reduce size of viral load simulation data (to 500 simulations) for CRAN
knitr::opts_chunk$set(root.dir = file.path(workDir,"saemix"))
setwd(file.path(workDir,"saemix"))

devtools::document(file.path(workDir,"saemix"))
roxygenise()
setwd(workDir)
system("R CMD build saemix")


testExamples<-TRUE
# Test examples
if(testExamples) system("R CMD check --as-cran --run-donttest saemix_3.0.tar.gz") else system("R CMD check --as-cran saemix_3.0.tar.gz")

if(FALSE)
  system("R CMD check --as-cran saemix_3.0.tar.gz")
```


#### Examples on CRAN

Examples with CPU or elapsed time > 5s
                 user system elapsed
rapi.saemix    32.510  0.092  32.602
cow.saemix     17.020  0.072  17.093
toenail.saemix 13.531  0.000  13.532
PD1.saemix     12.789  0.035  12.910
compare.saemix  6.331  0.016   6.347
theo.saemix     5.171  0.000   5.172

### Warnings

## Check

### Install package in development mode
  
```{r devModeInstall}
dev_mode() # development mode
install.packages(pkgs=file.path(workDir,"saemix_3.0.tar.gz"),repos=NULL)
library(saemix)
library(testthat)
```

### Running theopp example

Documentation not found

```{r helpTheo}
?theo.saemix
?saemix
```

Run on the theophylline example

```{r runTheo, echo=FALSE}
data(theo.saemix)

saemix.data<-saemixData(name.data=theo.saemix,header=TRUE,sep=" ",na=NA,
   name.group=c("Id"),name.predictors=c("Dose","Time"),
   name.response=c("Concentration"),name.covariates=c("Weight","Sex"),
   units=list(x="hr",y="mg/L", covariates=c("kg","-")), name.X="Time", verbose=FALSE)
print(saemix.data)

model1cpt<-function(psi,id,xidep) { 
	  dose<-xidep[,1]
	  tim<-xidep[,2]  
	  ka<-psi[id,1]
	  V<-psi[id,2]
	  CL<-psi[id,3]
	  k<-CL/V
	  ypred<-dose*ka/(V*(ka-k))*(exp(-k*tim)-exp(-ka*tim))
	  return(ypred)
}

saemix.model<-saemixModel(model=model1cpt,
  description="One-compartment model with first-order absorption", 
  psi0=matrix(c(1.,20,0.5,0.1,0,-0.01),ncol=3, byrow=TRUE,
  dimnames=list(NULL, c("ka","V","CL"))),transform.par=c(1,1,1),
  covariate.model=matrix(c(0,1,0,0,0,0),ncol=3,byrow=TRUE),fixed.estim=c(1,1,1),
  covariance.model=matrix(c(1,0,0,0,1,0,0,0,1),ncol=3,byrow=TRUE),
  omega.init=matrix(c(1,0,0,0,1,0,0,0,1),ncol=3,byrow=TRUE),error.model="constant", verbose=FALSE)
print(saemix.model)

saemix.fit<-saemix(model=saemix.model,
                   data=saemix.data,
                   control=list(seed=632545,directory="newtheo", save=FALSE,save.graphs=FALSE, warnings=FALSE))

# Prints a summary of the results
print(saemix.fit)

# Outputs the estimates of individual parameters
psi(saemix.fit)
```

Plot results:

```{r plotTheo, echo=FALSE}
# Shows some diagnostic plots to evaluate the fit
plot(saemix.fit)
plot(saemix.fit, plot.type="vpc")
plot(saemix.fit, plot.type="npde")
ynpde<-npdeSaemix(saemix.fit)
plot(saemix.fit, plot.type="individual")
```

```{r deactivateDevMode, echo=FALSE, hide=T}
dev_mode()
```

