---
title: "Saemix fits with different error models"
author: "Emmanuelle"
date: "30/04/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Directories
saemixDir<-"/home/eco/work/saemix/saemixextension"
progDir<-file.path(saemixDir,"R")
datDir<-file.path(saemixDir,"data")
testDir<-file.path(saemixDir,"testbelhal")
melDir<-file.path("/home/eco/work/saemix/bugs/melanie2004")

# Libraries
library(ggplot2)
library(MASS)

# Sourcing saemix functions
{
  source(file.path(progDir,"aaa_generics.R"))
  #source(file.path(progDir,"global.R"))
  source(file.path(progDir,"SaemixData.R"))
  source(file.path(progDir,"SaemixRes.R"))
  source(file.path(progDir,"SaemixModel.R"))
  source(file.path(progDir,"SaemixObject.R"))
  source(file.path(progDir,"main.R"))
  source(file.path(progDir,"func_aux.R"))
  source(file.path(progDir,"main_initialiseMainAlgo.R"))
  source(file.path(progDir,"main_estep.R"))
  source(file.path(progDir,"main_mstep.R"))
  source(file.path(progDir,"func_FIM.R"))
  source(file.path(progDir,"func_plots.R"))
  source(file.path(progDir,"func_distcond.R"))
  source(file.path(progDir,"func_simulations.R"))
  source(file.path(progDir,"compute_LL.R"))
  source(file.path(progDir,"func_estimParam.R"))
}
```

## Saemix fits with different error models

### Objective

Investigate why sometimes saemix doesn't converge and leads to outlandish estimates including variabilities of several thousands.

### Simulation using the same setting as Dubois et al. 2011

```{r simulData}
# Parameters
psi1<-c(1.5, 5, 0.04)
omega1<-diag(c(0.05, 0.0125, 0.05))
res1<-c(0.1,0.1)

# Model
model1cpt<-function(psi,id,xidep) {
  tim<-xidep[,1]
  dose<-xidep[,2]
  ka<-psi[id,1]
  V<-psi[id,2]
  CL<-psi[id,3]
  k<-CL/V
  ypred<-dose*ka/(V*(ka-k))*(exp(-k*tim)-exp(-ka*tim))
  return(ypred)
}

# Settings
N <- 50
tim <- c(0,0.25,0.5,1,2,3.5,5,7,9,12,24)
dose<-4

simdat<-data.frame(id=rep(1:N,each=length(tim)),time=rep(tim,N),dose=dose)

psipar<-do.call(rbind,rep(list(psi1),N))
for(i in 1:3) psipar[,i]<-psipar[,i]*exp(rnorm(N,mean=0,sd=sqrt(omega1[i,i])))
summary(psipar)
apply(psipar,2,sd)

ypred<-model1cpt(psipar,id=1:N,xidep=simdat[,2:3])
gpred<-error(ypred,res1,etype=rep(1,length(ypred)))
simdat$conc<-ypred+rnorm(length(ypred),mean=0,sd=gpred)

# Saemix data
saemix.data<-saemixData(name.data=simdat,header=TRUE,sep="",na=NA, name.group=c("id"), name.predictors=c("time","dose"),name.response=c("conc"))

simdat2<-simdat
simdat2$conc[simdat2$time==0]<-0
saemix.data2<-saemixData(name.data=simdat2,header=TRUE,sep="",na=NA, name.group=c("id"), name.predictors=c("time","dose"),name.response=c("conc"))
simdat3<-simdat2[simdat2$time>0,]
saemix.data3<-saemixData(name.data=simdat3,header=TRUE,sep="",na=NA, name.group=c("id"), name.predictors=c("time","dose"),name.response=c("conc"))

par(mfrow=c(1,2))
plot(simdat$time,simdat$conc,pch=20)
plot(simdat2$time,simdat2$conc,pch=20)

if(FALSE) {
  namfile<-file.path(melDir,"simulationsTlag.csv")
  simdat.mlx<-simdat
  simdat.mlx$dose[simdat.mlx$time>0]<-NA
  write.table(simdat.mlx,namfile, row.names=FALSE, quote=FALSE,na=".")
}
```


```{r fitSaemix.errorModels, echo=FALSE, results='hide',message=F, warning=F, include=F}
saemix.options<-list(nb.chains=3,seed=123456,save=FALSE, save.graphs=FALSE)

# Additive error
saemix.model1<-saemixModel(model=model1cpt,description="One compartment model", modeltype="structural", psi0=matrix(c(psi1,0,0,0),ncol=3,byrow=TRUE, dimnames=list(NULL, c("ka","V","CL"))), transform.par=c(1,1,1),  covariance.model=diag(3), error.model="constant")

# Proportional error
saemix.model2<-saemixModel(model=model1cpt,description="One compartment model", modeltype="structural", psi0=matrix(c(psi1,0,0,0),ncol=3,byrow=TRUE, dimnames=list(NULL, c("ka","V","CL"))), transform.par=c(1,1,1),  covariance.model=diag(3), error.model="proportional")
saemix.model2c<-saemixModel(model=model1cpt,description="One compartment model", modeltype="structural", psi0=matrix(c(psi1*.95,0,0,0),ncol=3,byrow=TRUE, dimnames=list(NULL, c("ka","V","CL"))), transform.par=c(1,1,1),  covariance.model=diag(3), error.model="proportional")

# Combined error
saemix.model3<-saemixModel(model=model1cpt,description="One compartment model", modeltype="structural", psi0=matrix(c(psi1,0,0,0),ncol=3,byrow=TRUE, dimnames=list(NULL, c("ka","V","CL"))), transform.par=c(1,1,1),  covariance.model=diag(3), error.model="combined")

fit1<-saemix(saemix.model1, saemix.data, saemix.options)
fit2<-saemix(saemix.model2, saemix.data, saemix.options)
fit2b<-saemix(saemix.model2, saemix.data2, saemix.options)
fit2c<-saemix(saemix.model2, saemix.data3, saemix.options)
fit3<-saemix(saemix.model3, saemix.data, saemix.options)
```

Showing results:
```{r fitResults}
print(fit1@results)
print(fit2@results)
print(fit2b@results)
print(fit3@results)


cat("Variances seem underestimated for V for constant and combined error models, and with modified data, and overestimated for all parameters for proportional error model:\n")
print(fit1@results@omega)
print(fit3@results@omega)
print(fit2b@results@omega)
print(fit2c@results@omega)

print(fit2@results@omega)

plot(fit1,plot.type="convergence")
plot(fit2,plot.type="convergence")

```


## Estimation with a proportional error model in Monolix

The runs fail to converge with a proportional error model.

- Model ka, V, Cl

Population parameter estimates
VALUE	STOCH. APPROX.
S.E.	R.S.E.(%)
Fixed Effects
ka_pop	2.87e+64	1.79e+67	6.23e+4
V_pop	2.58e-171	nan	nan
Cl_pop	1	nan	nan
Standard Deviation of the Random Effects
omega_ka	338	179	53
omega_V	204	27	13.3
omega_Cl	3.6e+56	nan	nan
Error Model Parameters
b	0.3	nan	nan

- Model ka, V, Cl, Tlag

Population parameter estimates
VALUE	STOCH. APPROX.
S.E.	R.S.E.(%)
Fixed Effects
Tlag_pop	1	nan	nan
ka_pop	1.08e+62	2.1e+65	1.94e+5
V_pop	1.34e-171	nan	nan
Cl_pop	1	nan	nan
Standard Deviation of the Random Effects
omega_Tlag	5.31e+57	nan	nan
omega_ka	346	538	155
omega_V	207	37.7	18.3
omega_Cl	1.63e+58	nan	nan
Error Model Parameters
b	0.3	nan	nan

# Problem with the error model, debugging step by step

Initialisation of the algorithm:

```{r saemix.initialisation}
saemix.model2<-saemixModel(model=model1cpt,description="One compartment model", modeltype="structural", psi0=matrix(c(psi1,0,0,0),ncol=3,byrow=TRUE, dimnames=list(NULL, c("ka","V","CL"))), transform.par=c(1,1,1),  covariance.model=diag(3), error.model="proportional")
saemix.options<-list(nb.chains=3,seed=123456,save=FALSE, save.graphs=FALSE)
saemix.data<-saemixData(name.data=simdat,header=TRUE,sep="",na=NA, name.group=c("id"), name.predictors=c("time","dose"),name.response=c("conc"))

  saemixObject<-new(Class="SaemixObject",data=saemix.data,model=saemix.model2,options=saemix.options)
  opt.warn<-getOption("warn")
  if(!saemixObject["options"]$warnings) options(warn=-1)

  saemix.options<-saemixObject["options"]
  saemix.model<-saemixObject["model"]
  saemix.data<-saemixObject["data"]
  saemix.data@ocov<-saemix.data@ocov[saemix.data@data[,"mdv"]==0,,drop=FALSE]
  saemix.data@data<-saemix.data@data[saemix.data@data[,"mdv"]==0,]
  saemix.data@ntot.obs<-dim(saemix.data@data)[1]

# Initialisation
  OLDRAND<-TRUE
  set.seed(saemix.options$seed)
  xinit<-initialiseMainAlgo(saemix.data,saemix.model,saemix.options)
  saemix.model<-xinit$saemix.model
  Dargs<-xinit$Dargs
  Uargs<-xinit$Uargs
  varList<-xinit$varList
  phiM<-xinit$phiM
  mean.phi<-xinit$mean.phi
  DYF<-xinit$DYF
  opt<-xinit$opt
  betas<-betas.ini<-xinit$betas
  fixed.psi<-xinit$fixedpsi.ini
  var.eta<-varList$diag.omega

  if (Dargs$modeltype=="structural"){
    theta0<-c(fixed.psi,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
    parpop<-matrix(data=0,nrow=(saemix.options$nbiter.tot+1),ncol=(Uargs$nb.parameters+length(Uargs$i1.omega2)+length(saemix.model["indx.res"])))
    colnames(parpop)<-c(saemix.model["name.modpar"], saemix.model["name.random"], saemix.model["name.sigma"][saemix.model["indx.res"]])
    allpar<-matrix(data=0,nrow=(saemix.options$nbiter.tot+1), ncol=(Uargs$nb.betas+length(Uargs$i1.omega2)+length(saemix.model["indx.res"])))
    colnames(allpar)<-c(saemix.model["name.fixed"],saemix.model["name.random"], saemix.model["name.sigma"][saemix.model["indx.res"]])
  } else{
    theta0<-c(fixed.psi,var.eta[Uargs$i1.omega2])
    parpop<-matrix(data=0,nrow=(saemix.options$nbiter.tot+1),ncol=(Uargs$nb.parameters+length(Uargs$i1.omega2)))
    colnames(parpop)<-c(saemix.model["name.modpar"], saemix.model["name.random"])
    allpar<-matrix(data=0,nrow=(saemix.options$nbiter.tot+1), ncol=(Uargs$nb.betas+length(Uargs$i1.omega2)))
    colnames(allpar)<-c(saemix.model["name.fixed"],saemix.model["name.random"])
  }
  
  parpop[1,]<-theta0
  allpar[1,]<-xinit$allpar0
  
  # using several Markov chains - only useful if passed back to main routine...
  # 	chdat<-new(Class="SaemixRepData",data=saemix.data, nb.chains=saemix.options$nb.chains)
  # 	NM<-chdat["NM"]
  # 	IdM<-chdat["dataM"]$IdM
  # 	yM<-chdat["dataM"]$yM
  # 	XM<-chdat["dataM"][,saemix.data["name.predictors"],drop=FALSE]
  
# List of sufficient statistics - change during call to stochasticApprox
  suffStat<-list(statphi1=0,statphi2=0,statphi3=0,statrese=0)
  phi<-array(data=0,dim=c(Dargs$N, Uargs$nb.parameters, saemix.options$nb.chains))

# structural model, check nb of parameters
  structural.model<-saemix.model["model"]
  #  nb.parameters<-saemix.model["nb.parameters"]
  
```

Burn-in iterations:

```{r saemix.burnin}
for (kiter in 1:saemix.options$nbiter.burn) { # Iterative portion of algorithm
# Burn-in - first loop useless
# E-step
  xmcmc<-estep(kiter, Uargs, Dargs, opt, structural.model, mean.phi, varList, DYF, phiM)
  varList<-xmcmc$varList
  DYF<-xmcmc$DYF
  phiM<-xmcmc$phiM

  # no M-step during burn-in phase
  allpar[(kiter+1),]<-allpar[kiter,]
  if(Dargs$modeltype=="structural") {
    theta<-c(fixed.psi,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
  } else{
    theta<-c(fixed.psi,var.eta[Uargs$i1.omega2])
  }
  parpop[(kiter+1),]<-theta
}
print(theta)
print(head(phiM))
print(varList$pres)
```

Something goes wrong in xstoch:

- sum(Dargs$yobs-fk)**2/cutoff(fk**2,.Machine$double.eps) explodes when predictions are very small (fk=0 or smaller than machine precision)
- changed code below to ignore the terms for which fk is too small in the summation

```{r saemix.explorationPhase1}
kiter<-saemix.options$nbiter.burn+1
	# E-step
  xmcmc<-estep(kiter, Uargs, Dargs, opt, structural.model, mean.phi, varList, DYF, phiM)
  varList<-xmcmc$varList
  DYF<-xmcmc$DYF
  phiM<-xmcmc$phiM

  # M-step and stochastic Approximation
  
#  xstoch<-mstep(kiter, Uargs, Dargs, opt, structural.model, DYF, phiM, varList, phi, betas, suffStat)

	# Update variances - TODO - check if here or elsewhere
	nb.etas<-length(varList$ind.eta)
	domega<-cutoff(mydiag(varList$omega[varList$ind.eta,varList$ind.eta]),.Machine$double.eps)
	omega.eta<-varList$omega[varList$ind.eta,varList$ind.eta,drop=FALSE]
	omega.eta<-omega.eta-mydiag(mydiag(varList$omega[varList$ind.eta,varList$ind.eta]))+mydiag(domega)
	#  print(varList$omega.eta)
	chol.omega<-try(chol(omega.eta))
	d1.omega<-Uargs$LCOV[,varList$ind.eta]%*%solve(omega.eta)
	d2.omega<-d1.omega%*%t(Uargs$LCOV[,varList$ind.eta])
	comega<-Uargs$COV2*d2.omega
	
	psiM<-transphi(phiM,Dargs$transform.par)
	fpred<-structural.model(psiM, Dargs$IdM, Dargs$XM)
  	for(ityp in Dargs$etype.exp) fpred[Dargs$XM$ytype==ityp]<-log(cutoff(fpred[Dargs$XM$ytype==ityp]))
#	if(Dargs$error.model=="exponential")
#		fpred<-log(cutoff(fpred))
	ff<-matrix(fpred,nrow=Dargs$nobs,ncol=Uargs$nchains)
	for(k in 1:Uargs$nchains) phi[,,k]<-phiM[((k-1)*Dargs$N+1):(k*Dargs$N),]
	# overall speed similar
	#    phi<-aperm(array(phiM,c(N,nchains,3)),c(1,3,2))
	stat1<-apply(phi[,varList$ind.eta,,drop=FALSE],c(1,2),sum) # sum on columns ind.eta of phi, across 3rd dimension
	stat2<-matrix(data=0,nrow=nb.etas,ncol=nb.etas)
	stat3<-apply(phi**2,c(1,2),sum) #  sum on phi**2, across 3rd dimension
	statr<-0
	for(k in 1:Uargs$nchains) {
		phik<-phi[,varList$ind.eta,k]
		stat2<-stat2+t(phik)%*%phik
		fk<-ff[,k]
		if(length(Dargs$error.model)==1) {
		  if(!is.na(match(Dargs$error.model,c("constant","exponential"))))
		    resk<-sum((Dargs$yobs-fk)**2) else {
		      if(Dargs$error.model=="proportional") {
		        idx.okpred<-which(fk>.Machine$double.eps)
		        vec<-(Dargs$yobs-fk)**2/cutoff(fk**2,.Machine$double.eps)
		        resk<-sum(vec[idx.okpred])
		        resk1<-sum(vec)
		        } else resk<-0
		    }
		} else resk<-0
		statr<-statr+resk
	}

print(resk)
print(resk1)
```


```{r saemix.explorationPhase1run, eval=FALSE}
kiter<-saemix.options$nbiter.burn+1
	# E-step
  xmcmc<-estep(kiter, Uargs, Dargs, opt, structural.model, mean.phi, varList, DYF, phiM)
  varList<-xmcmc$varList
  DYF<-xmcmc$DYF
  phiM<-xmcmc$phiM
  #  psiM<-transphi(phiM,saemix.model["transform.par"])
  
  #
  
  # M-step
#  if(opt$stepsize[kiter]>0) {
############# Stochastic Approximation
  	xstoch<-mstep(kiter, Uargs, Dargs, opt, structural.model, DYF, phiM, varList, phi, betas, suffStat)
  	varList<-xstoch$varList
  	mean.phi<-xstoch$mean.phi
  	phi<-xstoch$phi
  	betas<-xstoch$betas
  	suffStat<-xstoch$suffStat
  	
  	beta.I<-betas[Uargs$indx.betaI]
  	fixed.psi<-transphi(matrix(beta.I,nrow=1),saemix.model["transform.par"])
  	betaC<-betas[Uargs$indx.betaC]
  	var.eta<-mydiag(varList$omega)
  	l1<-betas.ini
  	l1[Uargs$indx.betaI]<-fixed.psi
  	l1[Uargs$indx.betaC]<-betaC

    if(Dargs$modeltype=="structural") {
      allpar[(kiter+1),]<-c(l1,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
    } else{
      allpar[(kiter+1),]<-c(l1,var.eta[Uargs$i1.omega2])
    }

  # } else { #end of loop on if (stepsize[kiter]>0)
  #   allpar[(kiter+1),]<-allpar[kiter,]
  # }
   if(Dargs$modeltype=="structural") {
      theta<-c(fixed.psi,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
    } else{
      theta<-c(fixed.psi,var.eta[Uargs$i1.omega2])
    }
  parpop[(kiter+1),]<-theta

print(theta)
print(head(phiM))
print(varList$pres)

```


```{r FIMforfit2b, eval=FALSE}
saemixObject<-fit2b

```


```{r saemix.explorationPhase, eval=FALSE}
kiter<-saemix.options$nbiter.burn+1

for (kiter in (saemix.options$nbiter.burn+1):(saemix.options$nbiter.sa)) { # 
# Burn-in - resetting sufficient statistics
  if(opt$flag.fmin && kiter==saemix.options$nbiter.sa) {
    cat("Inside first loop, kiter=",kiter,":\n")
  	Uargs$COV1<-Uargs$COV[,Uargs$ind.fix11]
  	ind.prov<-!(varList$ind.eta %in% Uargs$i0.omega2)
  	varList$domega2<-varList$domega2[ind.prov,ind.prov,drop=FALSE] # keep in domega2 only indices of parameters with IIV
  	varList$ind0.eta<-Uargs$i0.omega2
  	varList$ind.eta<-1:(Uargs$nb.parameters)  	
  	if(length(varList$ind0.eta)>0) varList$ind.eta<-varList$ind.eta[!(varList$ind.eta %in% varList$ind0.eta)] # update ind.eta, now only parameters with IIV
  	Uargs$nb.etas<-length(varList$ind.eta)
  	suffStat$statphi1<-0
  	suffStat$statphi2<-0
  	suffStat$statphi3<-0
  }

	# E-step
  xmcmc<-estep(kiter, Uargs, Dargs, opt, structural.model, mean.phi, varList, DYF, phiM)
  varList<-xmcmc$varList
  DYF<-xmcmc$DYF
  phiM<-xmcmc$phiM
  #  psiM<-transphi(phiM,saemix.model["transform.par"])
  
  # M-step
  if(opt$stepsize[kiter]>0) {
############# Stochastic Approximation
  	xstoch<-mstep(kiter, Uargs, Dargs, opt, structural.model, DYF, phiM, varList, phi, betas, suffStat)
  	varList<-xstoch$varList
  	mean.phi<-xstoch$mean.phi
  	phi<-xstoch$phi
  	betas<-xstoch$betas
  	suffStat<-xstoch$suffStat
  	
  	beta.I<-betas[Uargs$indx.betaI]
  	fixed.psi<-transphi(matrix(beta.I,nrow=1),saemix.model["transform.par"])
  	betaC<-betas[Uargs$indx.betaC]
  	var.eta<-mydiag(varList$omega)
  	l1<-betas.ini
  	l1[Uargs$indx.betaI]<-fixed.psi
  	l1[Uargs$indx.betaC]<-betaC

    if(Dargs$modeltype=="structural") {
      allpar[(kiter+1),]<-c(l1,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
    } else{
      allpar[(kiter+1),]<-c(l1,var.eta[Uargs$i1.omega2])
    }

  } else { #end of loop on if (stepsize[kiter]>0)
    allpar[(kiter+1),]<-allpar[kiter,]
  }
   if(Dargs$modeltype=="structural") {
      theta<-c(fixed.psi,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
    } else{
      theta<-c(fixed.psi,var.eta[Uargs$i1.omega2])
    }
  parpop[(kiter+1),]<-theta

# End of loop on kiter
}
print(theta)
print(head(phiM))
print(varList$pres)

if(FALSE) {
   namfile<-file.path(melDir,"simulationsTlag.csv")
   dat1<-read.table(namfile,header=T,na=".")
   dat1$conc[dat1$time==0]<-NA
   namfile2<-file.path(melDir,"simulationsTlag_conc0.csv")
   write.table(dat1,namfile2, row.names=FALSE, quote=FALSE,na=".")
}

```

## Dataset failing with a combined error model

```{r combinedErrorPb}
load(file.path(melDir,"combinedError","saemix_simu"))

## Structural model : One compartment, first-order absorption, linear elimination, noTlag ----
model1cpt_firstorder=function(psi,id,xidep) { 
  dose=4
  tim=xidep[,1]  
  ka=psi[id,1]
  V=psi[id,2]
  CL=psi[id,3]
  k=CL/V
  ypred=dose*ka/(V*(ka-k))*(exp(-k*tim)-exp(-ka*tim))
  return(ypred)
}

## Statistical model
saemix_model=saemixModel(model=model1cpt_firstorder,
                           description="One-compartment model with first-order absorption, no time lag, no F parameter",
                           psi0=matrix(c(1.3,0.55,0.0415),
                                       ncol=3,byrow=TRUE, 
                                       dimnames=list(NULL, c("ka","V","CL"))),
                           covariate.model = c(1,1,1),
                           transform.par=c(1,1,1),
                           error.model="combined")

## Model ajustment
saemix_options=list(seed=632545,save=FALSE,save.graphs=FALSE,nb.chains=3,nbiter.saemix=c(200,100))

saemix_options=list(seed=152093,save=FALSE,save.graphs=FALSE,nb.chains=3,nbiter.saemix=c(200,100))
saemix_fit=saemix(saemix_model,saemix_simu,saemix_options)

```

```{r combinedErrorPb.investigation1}
saemix_options=list(seed=632545,save=FALSE,save.graphs=FALSE,nb.chains=3,nbiter.saemix=c(200,100))

saemixObject<-new(Class="SaemixObject",data=saemix_simu,model=saemix_model,options=saemix_options)

saemix.options<-saemixObject["options"]
saemix.model<-saemixObject["model"]
saemix.data<-saemixObject["data"]
saemix.data@ocov<-saemix.data@ocov[saemix.data@data[,"mdv"]==0,,drop=FALSE]
saemix.data@data<-saemix.data@data[saemix.data@data[,"mdv"]==0,]
saemix.data@ntot.obs<-dim(saemix.data@data)[1]

# Initialisation
OLDRAND<-TRUE
set.seed(saemix.options$seed)
xinit<-initialiseMainAlgo(saemix.data,saemix.model,saemix.options)
saemix.model<-xinit$saemix.model
Dargs<-xinit$Dargs
Uargs<-xinit$Uargs
varList<-xinit$varList
phiM<-xinit$phiM
mean.phi<-xinit$mean.phi
DYF<-xinit$DYF
opt<-xinit$opt
betas<-betas.ini<-xinit$betas
fixed.psi<-xinit$fixedpsi.ini
var.eta<-varList$diag.omega

if (Dargs$modeltype=="structural"){
  theta0<-c(fixed.psi,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
  parpop<-matrix(data=0,nrow=(saemix.options$nbiter.tot+1),ncol=(Uargs$nb.parameters+length(Uargs$i1.omega2)+length(saemix.model["indx.res"])))
  colnames(parpop)<-c(saemix.model["name.modpar"], saemix.model["name.random"], saemix.model["name.sigma"][saemix.model["indx.res"]])
  allpar<-matrix(data=0,nrow=(saemix.options$nbiter.tot+1), ncol=(Uargs$nb.betas+length(Uargs$i1.omega2)+length(saemix.model["indx.res"])))
  colnames(allpar)<-c(saemix.model["name.fixed"],saemix.model["name.random"], saemix.model["name.sigma"][saemix.model["indx.res"]])
} else{
  theta0<-c(fixed.psi,var.eta[Uargs$i1.omega2])
  parpop<-matrix(data=0,nrow=(saemix.options$nbiter.tot+1),ncol=(Uargs$nb.parameters+length(Uargs$i1.omega2)))
  colnames(parpop)<-c(saemix.model["name.modpar"], saemix.model["name.random"])
  allpar<-matrix(data=0,nrow=(saemix.options$nbiter.tot+1), ncol=(Uargs$nb.betas+length(Uargs$i1.omega2)))
  colnames(allpar)<-c(saemix.model["name.fixed"],saemix.model["name.random"])
}

parpop[1,]<-theta0
allpar[1,]<-xinit$allpar0
# List of sufficient statistics - change during call to stochasticApprox
suffStat<-list(statphi1=0,statphi2=0,statphi3=0,statrese=0)
phi<-array(data=0,dim=c(Dargs$N, Uargs$nb.parameters, saemix.options$nb.chains))

# structural model, check nb of parameters
structural.model<-saemix.model["model"]
#  nb.parameters<-saemix.model["nb.parameters"]

# Burn-in
cat("\nBurn-in phase\n")
for (kiter in 1:saemix.options$nbiter.burn) { # Iterative portion of algorithm
# Burn-in - first loop useless
# E-step
  xmcmc<-estep(kiter, Uargs, Dargs, opt, structural.model, mean.phi, varList, DYF, phiM)
  varList<-xmcmc$varList
  DYF<-xmcmc$DYF
  phiM<-xmcmc$phiM

  # no M-step during burn-in phase
  allpar[(kiter+1),]<-allpar[kiter,]
  if(Dargs$modeltype=="structural") {
    theta<-c(fixed.psi,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
  } else{
    theta<-c(fixed.psi,var.eta[Uargs$i1.omega2])
  }
  parpop[(kiter+1),]<-theta
}
print(theta)
print(head(phiM))
print(varList$pres)

# Simulated annealing phase (K1)
kiter<-saemix.options$nbiter.burn+1
cat("\nRunning K1 iterations\n")
for (kiter in (saemix.options$nbiter.burn+1):(saemix.options$nbiter.sa)) { # 
# Burn-in - resetting sufficient statistics
  if(opt$flag.fmin && kiter==saemix.options$nbiter.sa) {
    cat("Inside first loop, kiter=",kiter,":\n")
  	Uargs$COV1<-Uargs$COV[,Uargs$ind.fix11]
  	ind.prov<-!(varList$ind.eta %in% Uargs$i0.omega2)
  	varList$domega2<-varList$domega2[ind.prov,ind.prov,drop=FALSE] # keep in domega2 only indices of parameters with IIV
  	varList$ind0.eta<-Uargs$i0.omega2
  	varList$ind.eta<-1:(Uargs$nb.parameters)  	
  	if(length(varList$ind0.eta)>0) varList$ind.eta<-varList$ind.eta[!(varList$ind.eta %in% varList$ind0.eta)] # update ind.eta, now only parameters with IIV
  	Uargs$nb.etas<-length(varList$ind.eta)
  	suffStat$statphi1<-0
  	suffStat$statphi2<-0
  	suffStat$statphi3<-0
  }

	# E-step
  xmcmc<-estep(kiter, Uargs, Dargs, opt, structural.model, mean.phi, varList, DYF, phiM)
  varList<-xmcmc$varList
  DYF<-xmcmc$DYF
  phiM<-xmcmc$phiM
  #  psiM<-transphi(phiM,saemix.model["transform.par"])
  
  # M-step
  if(opt$stepsize[kiter]>0) {
############# Stochastic Approximation
  	xstoch<-mstep(kiter, Uargs, Dargs, opt, structural.model, DYF, phiM, varList, phi, betas, suffStat)
  	varList<-xstoch$varList
  	mean.phi<-xstoch$mean.phi
  	phi<-xstoch$phi
  	betas<-xstoch$betas
  	suffStat<-xstoch$suffStat
  	
  	beta.I<-betas[Uargs$indx.betaI]
  	fixed.psi<-transphi(matrix(beta.I,nrow=1),saemix.model["transform.par"])
  	betaC<-betas[Uargs$indx.betaC]
  	var.eta<-mydiag(varList$omega)
  	l1<-betas.ini
  	l1[Uargs$indx.betaI]<-fixed.psi
  	l1[Uargs$indx.betaC]<-betaC

    if(Dargs$modeltype=="structural") {
      allpar[(kiter+1),]<-c(l1,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
    } else{
      allpar[(kiter+1),]<-c(l1,var.eta[Uargs$i1.omega2])
    }

  } else { #end of loop on if (stepsize[kiter]>0)
    allpar[(kiter+1),]<-allpar[kiter,]
  }
   if(Dargs$modeltype=="structural") {
      theta<-c(fixed.psi,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
    } else{
      theta<-c(fixed.psi,var.eta[Uargs$i1.omega2])
    }
  parpop[(kiter+1),]<-theta

# End of loop on kiter
}
cat("kiter=",kiter,"\n")
print(theta)
print(head(phiM))
print(varList$pres)
print(theta)

```


```{r combinedErrorPb.investigation2}
for (kiter in (saemix.options$nbiter.sa+1):(saemix.options$nbiter.sa+1)) { # Iterative portion of algorithm
	# E-step
  xmcmc<-estep(kiter, Uargs, Dargs, opt, structural.model, mean.phi, varList, DYF, phiM)
  varList<-xmcmc$varList
  DYF<-xmcmc$DYF
  phiM<-xmcmc$phiM
  #  psiM<-transphi(phiM,saemix.model["transform.par"])
  
  # M-step
  if(opt$stepsize[kiter]>0) {
############# Stochastic Approximation
  	xstoch<-mstep(kiter, Uargs, Dargs, opt, structural.model, DYF, phiM, varList, phi, betas, suffStat)
  	varList<-xstoch$varList
  	mean.phi<-xstoch$mean.phi
  	phi<-xstoch$phi
  	betas<-xstoch$betas
  	suffStat<-xstoch$suffStat
  	
  	beta.I<-betas[Uargs$indx.betaI]
  	fixed.psi<-transphi(matrix(beta.I,nrow=1),saemix.model["transform.par"])
  	betaC<-betas[Uargs$indx.betaC]
  	var.eta<-mydiag(varList$omega)
  	l1<-betas.ini
  	l1[Uargs$indx.betaI]<-fixed.psi
  	l1[Uargs$indx.betaC]<-betaC

    if(Dargs$modeltype=="structural") {
      allpar[(kiter+1),]<-c(l1,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
    } else{
      allpar[(kiter+1),]<-c(l1,var.eta[Uargs$i1.omega2])
    }

  } else { #end of loop on if (stepsize[kiter]>0)
    allpar[(kiter+1),]<-allpar[kiter,]
  }
   if(Dargs$modeltype=="structural") {
      theta<-c(fixed.psi,var.eta[Uargs$i1.omega2],varList$pres[Uargs$ind.res])
    } else{
      theta<-c(fixed.psi,var.eta[Uargs$i1.omega2])
    }
  parpop[(kiter+1),]<-theta

# End of loop on kiter
}
cat("kiter=",kiter,"\n")
print(theta)
print(head(phiM))
print(varList$pres)
print(theta)

print(xmcmc$varList$pres)
print(xstoch$varList$pres)

```
