---
title: "Changing classes in saemix - introducing IOV"
author: "Emmanuelle Comets"
date: "2026-02-13"
output:
  pdf_document: default
  html_document: default
---

# Objective

```{r setup, include=FALSE}
# Setting working directory
require("knitr")
knitr::opts_chunk$set(echo = TRUE)

# Libraries
library(rlang)
library(ggplot2) 
library(RColorBrewer) 
library(gridExtra)
library(viridis)
library(testthat)

#library(dplyr)
# library(Cairo)

# Working directory
saemixDir<-"/home/eco/work/saemix/saemixextension"
workDir <-getwd()
setwd(workDir)
knitr::opts_chunk$set(root.dir = workDir)

# Load function generics (same for both versions)
progDir.current<-file.path(saemixDir,"R")
progDir.next<-file.path(saemixDir,"Rext")
source(file.path(progDir.current,"aaa_generics.R"))

# directory for new data
datDir<-file.path(saemixDir,"data40")

```

# Simulation of IOV with theophylline PK

- Features
  - theophylline PK
  - covariates
    - sex, weight changing with subject
    - CRCL, comedication changing with occasion
  - design
    - N=100 subjects
    - 1 to 3 occasion per subject (random)
    - sampling times for each occasion: 0.25, 1.5, 3.35, 12 and 24 h (sparse design in Dubois et al.)
    - dose: D=4mg
    - time between two occasions: 4 weeks (assume washout so no drug left)
- Model
  - PK response: 1 cpt model with absorption

    $$ Cpred = \frac{D}{V} \frac{k_a}{(k_a-CL/V)} \left( e^{- CL/V t} - e^{-k_a t} \right)$$
  - log-normal parameter distribution
  - population parameter values (Dubois et al.)
    - ka=1.48 h-1
    - CL=0.04 L/h
    - V=0.48 L
  - variabilities
    - IIV 
      - 30% on all parameters
      - correlation between CL and V (0.7)
    - IOV
      - IOV on CL only with 25% IOV
    - different from Dubois, where 2 scenarios, no correlation and no covariate effect
      - low IIV: 0.2 on ka and CL, 0.1 on V; half of IIV for IOV
      - regular BSV: 0.5 on ka, V, CL and 0.15 on IOV
    - combined residual error model with a=0.1 and b=0.1 (low IIV scenario from Dubois)
  - covariate models
    - subject-level:
      - weight: allometry on V and CL
      - gender: V decreased by 30% in women
    - occasion level
      - CRCL: correlation with CL ($\beta_{CRCL, CL}=0.5$)
      - comedication: increase in CL by 30\%
- Not included in the simulation
  - NA values in the covariates
  - LOQ values

### Read datasets

- Theophylline without IOV on CL
  - but small differences due to covariates for the second simulation

```{r readSimulatedTheoDataIIV}
# Dataset without covariates
theonocov <- read.csv(file=file.path(datDir,"theoSimul_nocov.csv"), header=T, sep=",")
psinocov <- read.csv(file=file.path(datDir,"theoSimul_psinocov.csv"), header=T, sep=",")

# Dataset with covariates (different psi across occasions but only due to covariates)
theocov <- read.csv(file=file.path(datDir,"theoSimul_cov.csv"), header=T, sep=",")
psicov <- read.csv(file=file.path(datDir,"theoSimul_psicov.csv"), header=T, sep=",")
```

### SaemixData object (new class)

- done
  - read data
  - define variability levels
  - associate indices at the different levels
- **ToDo**
  - need to modify SaemixRepData so that it takes on the highest level of variability for matching
- **ToDo after matching with model**
  - check consistency with covariates that are in the model (keep in @data for graphs, adjust model with warnings if covariates not found)
  - check consistency of variability levels with model (presence of levels of variability)
  - check consistency of covariate models at different levels of variability
    - warning if inconsistency (eg covariate at IIV level but varying across occasions, covariate at IOV level but in fact related to IIV)
  - extract matrix of covariates at each variability level
    - add to SaemixData object ? (probably, but not sure whether to keep only the covariates specified in the model... or maybe add also a slot at lowest variability level with the all the covariate values)
  - redefine SaemixRepData if the levels of variability have changed

```{r newSaemixData, results='hide', echo=FALSE, warnings=FALSE}
source(file.path(progDir.next,"SaemixData.R"))
source(file.path(progDir.next,"SaemixData-printmethods.R"))
source(file.path(progDir.next,"SaemixData-covariatemethods.R"))
source(file.path(progDir.next,"SaemixData-methods.R"))
source(file.path(progDir.next,"SaemixData-plotmethods.R"))

theodat.iov <-  saemixData(name.data=file.path(datDir,"theoSimul_nocov.csv"), header=T,na=".", sep=",", varlevel=c("id","occ"),name.predictors=c("tim", "dose"),name.response=c("ysim"), unit=list(x="hr",y="mg/L"), verbose=TRUE)

xdata <- theodat.iov@data
occ.labs <-paste("Occasion",1:3)
names(occ.labs) <- c("1", "2", "3")

ggplot(xdata, aes(x=tim, y=ysim, group=id, colour=as.factor(occ))) + geom_point() + geom_line(data = subset(xdata)) + xlab("Time (hr)") + ylab("Concentrations (mg/L)")  + scale_color_viridis("Occasion",discrete = TRUE) +  facet_wrap(.~occ, labeller = labeller(occ=occ.labs, .multi_line=FALSE))

# No covariates, no iov
theodat.iiv <-  saemixData(name.data=file.path(datDir,"theoSimul_nocov.csv"), header=T,na=".", sep=",", varlevel=c("id"),name.predictors=c("tim", "dose"),name.response=c("ysim"), unit=list(x="hr",y="mg/L"), verbose=TRUE)
cat("Not ordered in the same way when occasion is ignored (sorted by time, otherwise sorted by occ then time\n")
summary(theodat.iiv@data$ysim - theodat.iov@data$ysim)
head(theodat.iov@data[theodat.iov@data$id==2,],10)
head(theodat.iiv@data[theodat.iiv@data$id==2,],10)

# With covariates and IOV
theocovdat <-  saemixData(name.data=file.path(datDir,"theoSimul_cov.csv"), header=T,na=".", sep=",", varlevel=c("id","occ"),name.predictors=c("tim", "dose"),name.response=c("ysim"), name.covariates=c("wt","sex","crcl","comed"), covariate.varlevel = c(rep("id",2),rep("occ",2)), unit=list(x="hr",y="mg/L",covariates=c("kg","-","g/dL","-")), verbose=TRUE)

# ToDo: 
## Done: automate: done is readSaemix, calling extractVarlevelCovariates
### ToDo: check variability levels in data object 
### automatically ascribe variability levels (currently goes to lowest variability level, ie occ here if we don't specify covariate.varlevel)
## ToDo: match data and model, matching covariates in data with popmodel matrices in saemix.model
if(FALSE){
  idx.iiv <-which(!duplicated(theocovdat@data$id))
  Mcov.iiv <- cbind(pop=1,theocovdat@data[idx.iiv,c("wt","sex")])
  ilab<-paste0(theocovdat@data$id,"-",theocovdat@data$occ)
  # Here no mu_occ (otherwise also need a column of 1's)
  idx.iov <-which(!duplicated(ilab))
  Mcov.iov <- theocovdat@data[idx.iov,c("crcl","comed")]
  Mcovar <- list(Mcov.iiv, Mcov.iov)
  names(Mcovar) <- theocovdat@varlevel
}
Mcovar<-theocovdat@var.covmat
head(Mcovar$id)
head(Mcovar$occ)

head(theodat.iov@data,12)
head(theodat.iiv@data,12)

cat("Expect 0:\n")
print(summary(theodat.iov@var.index[[2]]-theodat.iov@data$index))
print(summary(theodat.iiv@var.index[[1]]-theodat.iiv@data$index))
```

### SaemixModel object (new class)


```{r newSaemixModel, results='hide', echo=FALSE, warnings=FALSE}
# Load subclasses for outcome, parameter, error model, variability model, individual, population and statistical models
source(file.path(progDir.next,"SaemixErrorModel.R"))
source(file.path(progDir.next,"SaemixOutcome.R"))
source(file.path(progDir.next,"SaemixParameter.R"))
source(file.path(progDir.next,"SaemixParameter-methods.R"))
source(file.path(progDir.next,"SaemixVarModel.R"))
source(file.path(progDir.next,"SaemixVarModel-methods.R"))
source(file.path(progDir.next,"SaemixPopModel.R"))
source(file.path(progDir.next,"SaemixIndivModel.R"))
source(file.path(progDir.next,"SaemixModel.R"))
source(file.path(progDir.next,"SaemixModel-methods.R"))

# model function
theo1cpt <- function(psi,id,xidep) {
  tim<-xidep[,1]
  dose<-xidep[,2]
  ka<-psi[id,1]
  V<-psi[id,2]
  CL<-psi[id,3]
  ypred <- dose*ka/V/(ka-CL/V)*(exp(-CL/V*tim) - exp(-ka*tim))
  return(ypred)
}

############### New definition, no IOV in the model
param.noiov <- list(ka=saemixParam(mu.init=1.5),
                vd=saemixParam(mu.init=0.5, sd.init=0.7,varlevel=c("iiv")),
                cl=saemixParam(name="cl",mu.init=0.04, varlevel=c("iiv"), sd.init=c(0.6), corr = list(iiv=c("vd")))
                )
theomodel.iiv <- saemixModel(model=theo1cpt, parameters=param.noiov)

############### New definition, with IOV and covariates in the model
# Model without covariates
param.iov <- list(ka=saemixParam(mu.init=1.5),
                vd=saemixParam(mu.init=0.5, sd.init=0.7,varlevel=c("iiv","iov")),
                cl=saemixParam(name="cl",mu.init=0.04, varlevel=c("iiv","iov"), sd.init=c(0.6,0.3), corr = list(iiv=c("vd")))
)
theomodel.iov <- saemixModel(model=theo1cpt, parameters=param.iov)

# Model with covariates
param.cov <- list(ka=saemixParam(mu.init=1.5),
                vd=saemixParam(mu.init=0.5, sd.init=0.7, covariate=c("wt","comed"), covariate.init=c(1,0), covariate.estim=c("fixed","estimated"), covariate.varlevel=c("iiv","iov")),  
                cl=saemixParam(name="cl",mu.init=2, varlevel=c("iiv","iov"), sd.init=c(0.6,0.3), corr = list(iiv=c("vd")), covariate=c("wt","sex","crcl"), covariate.init=c(0.75,0,0), covariate.estim=c("fixed","estimated","estimated"), corr.init=list(iiv=c(-0.5,0.7), iov=0.7), covariate.varlevel=c("iiv","iiv","iov"))
)
theomodel.coviov <- saemixModel(model=theo1cpt, parameters=param.cov)
theomodel.coviov@popmodel

# Extraction of parameters
```

### Initialisation

- Create lists needed by the algorithm
  - Margs: model-related elements (passed on to functions, unchanged)
    - structural model
    - residual error models/discrete nature of outcomes
  - Dargs: data-related elements (passed on to functions, unchanged)
    - observed data (yobs) and predictors
    - indices of non-zero elements in DYF
    -...
  - Vargs
    - omega matrices
    - indices
- compute.LLy: computes the log-likelihood given individual parameters phiM
  - uses data in Dargs (initialised in createStructData)
  - uses model features in Margs (initialised in createStructModel)
- **ToDo M2 Lille**
  - check if extends to IOV (should not change the computation)
    - seems to work, but need to check
  - check if extends to covariates (should not change the computation)
    - seems to work, but need to check
  - check if extends to multiple chains (need to expand phiM)
- **ToDo M2 Lille**
  - matching functions
    - check matching between
    - create design matrices to hold the individual covariate structures
  - complete createStructData to extract covariate matrices
- **ToDo** Eco
  - add initialisation of Vargs (variables updated during the fit) to createStructModel (variables extracted initially from the model)
  - for the moment specific function *createVarlist()*, see if need to regroup

```{r initialLists, results='hide', echo=FALSE, warnings=FALSE}
# Computational functions
source(file.path(progDir.next,"func_aux.R"))
# Initialise functions 
source(file.path(progDir.next,"func_initialise.R"))

# IIV only
# Dargs from theodat.iiv (no covariates)
dlist <- createStructData(theodat.iiv, nb.chains=1)
Dargs.iiv <- dlist$Dargs
DYF.iiv <- dlist$DYF
cat("Expect 0:\n")
print(summary(Dargs.iiv$IdM-theodat.iiv@var.index[[1]]))
# Margs from theomodel.iiv (no covariates, only IIV)
mlist <- createStructModel(theomodel.iiv)
Margs.iiv <- mlist$Margs

# IOV + IIV
# Dargs from theodat.iov (no covariates)
dlist <- createStructData(theodat.iov, nb.chains=1)
Dargs.iov <- dlist$Dargs
DYF.iov <- dlist$DYF
cat("Expect 0:\n")
print(summary(Dargs.iov$IdM-theodat.iov@var.index[[2]]))
# Margs from theomodel.iiv (no covariates, only IIV)
mlist <- createStructModel(theomodel.iiv)
Margs.iov <- mlist$Margs


```


### Computational functions

- model predictions
  - need to use the second level of indices (var.index[[2]]) to match the occasion-level parameters (extracted in Dargs)
  - recovers the same values as ypred (without residual error)
- error
  - note how this function is used
  - see also testthat_auxFunc.R
- ssq
  - compare yobs with fpred
  - see also testthat_auxFunc.R
- computation of likelihood associated to observations
  - input
    - phiM (individual parameters on the eta scale, as a matrix)
    - lists Dargs (elements related to data) and Margs (elements related to model): created by specific functions (see above)
    - matrix DYF also created in the same function
  - **ToDo:** check (compute fpred and gpred by hand and compare for each subject)

```{r auxFunctions, results='hide', echo=FALSE, warnings=FALSE}
# Using Dargs to predict 
psiM <- as.matrix(psinocov[,c(2:4)])
#fpred <- theomodel@model(psiM, theodat.iov@var.index[[2]], Dargs$XM)
fpred <- theomodel.iov@model(psiM, Dargs.iov$IdM, Dargs.iov$XM)
expect_true(max(fpred-theonocov$ypred)<10-6) # almost zero

# Error model
gpred<-error(fpred,list(theomodel.iov@outcome[[1]]@error.model), theodat.iov@data$ytype)
expect_equal(min(gpred),1)
expect_equal(max(gpred),1)
summary(gpred) # should be 1 (constant variance with parameter initialised at 1)

# Sum of squares (see also the associated testtaht (WiP))
xcal <- ssq(theomodel.iov@outcome[[1]]@error.model@param, Dargs.iov$yobs, fpred, theomodel.iov@outcome[[1]]@error.model)
xcal

# Computation of likelihood associated to observations - IIV
psiM.iiv <- as.matrix(psinocov[!duplicated(psinocov$id),c(2:4)])
expect_equal(dim(psiM.iiv)[1],dim(DYF.iiv)[2])
phiM.iiv <- transformPar(psiM.iiv,Margs.iiv$invtransform)
LLy.iiv <- compute.LLy(phiM.iiv, Margs.iiv, Dargs.iiv, DYF.iiv) 
sum(LLy.iiv)

# Computation of likelihood associated to observations - IIV + IOV
psiM.iov <- as.matrix(psinocov[,c(2:4)])
phiM.iov <- transformPar(psiM.iov,Margs.iov$invtransform)
LLy.iov <- compute.LLy(phiM.iov, Margs.iov, Dargs.iov, DYF.iov) 
sum(LLy.iov)

cat("Quick check: model with IOV should have better LL than the model without, here ")
if(expect_true(sum(LLy.iiv)>sum(LLy.iov))) cat("TRUE \n")
# Should be true for all subjects 
## same dataset, one model has 'better' parameters adjusted to occasions, but the order of observations is differnet in both datasets so can't compare element to element ? (should still be the same subject order...
# data.frame(LLy.iiv[which(LLy.iiv<LLy.iov)],LLy.iov[which(LLy.iiv<LLy.iov)])
```

### Simulation (E-step)

#### Simulations in the extended matrix

- simulation at the level of the full vector of individual parameters
  - not at the eta level
  - uses the extended matrix (denoted $\Gamma$ in the Panhard and Samson paper)
- list Vargs holding omega matrices (created by *createVarList()*)
  - WiP: will probably need additional elements previously contained in list varList
  - note: elements of the residual error now in error.model, either update them there or keep track of them in Vargs
- **ToDo** simulation
  - simulation of etas
    - simulate in extended matrix
    - here use svd decomposition as cholesky doesn't work (the extended matrix isn't invertible)
    - **ToDo**: check if the simulation is correct
  - extend to different number of occasions across subjects
    - possible solution: simulate as a block then keep only occasions present for each subject
    - may need a second series of indices as lists in ind.ioM
  - extend to covariates
- **ToDo** E-step
  - simulation in each kernel
  - kernel with block structure (simulation of a subset of parameters): not obvious how to do
    - think about solutions (repeated blocks across the nocc ?)

```{r simulEta, results='hide', echo=FALSE, warnings=FALSE}
# Model without covariates,  IIV + IOV
smx.model <- theomodel.iov
smx.data <- theodat.iov

dlist <- createStructData(smx.data, nb.chains=1)
Dargs <- dlist$Dargs
DYF <- dlist$DYF

# Margs from theomodel.iov
mlist <- createStructModel(smx.model)
Margs <- mlist$Margs

vlist <- createVarlist(smx.model, smx.data)
Vargs <- vlist$Vargs
omega.sim <- vlist$Vargs$omega.sim
ind.eta <- vlist$ind.eta
nb.etas <- length(ind.eta)
ind.eta.ext <- vlist$ind.eta.ext
nb.etas.sim <- length(ind.eta.ext)
domega <- cutoff(mydiag(omega.sim[ind.eta.ext,ind.eta.ext]),.Machine$double.eps)
omega.eta <- omega.sim[ind.eta.ext,ind.eta.ext, drop=FALSE]
omega.eta <- omega.eta-mydiag(mydiag(omega.sim[ind.eta.ext,ind.eta.ext]))+mydiag(domega)

# Can't really simulate like in the E-step without IOV because the blocks make the matrix not invertible
if(FALSE) {
  chol.omega<-try(chol(omega.eta))
	somega<-solve(omega.eta)
  nb.etas.sim<-length(ind.eta.ext)
	etaMc<-try(matrix(rnorm(Dargs$NM*nb.etas.sim),ncol=nb.etas.sim)%*%chol.omega)
}

# chol.omega<-try(chol(omega.eta))
# Cholesky decomposition fails with extended matrix => using svd (alternative approach implemented in rmvtnorm)
# 2-3 times slower
svd.omega<-svd(omega.eta)
rmat.omega <- t(svd.omega$v %*% (t(svd.omega$u) * sqrt(pmax(svd.omega$d, 0))))
etaMc<-matrix(rnorm(Dargs$NM*nb.etas.sim),ncol=nb.etas.sim)%*%rmat.omega
head(etaMc)
## now need to compute mean.phi for phiM then
## U.y <- compute.LLy(phiM, Margs, Dargs, DYF) 

```


#### Trying to compute the mean of simulated etas over all occasions

- Simulation of etas $\phi_{ik}$ in the extended matrix (nb.etas x nocc.max)
- Storage structure for etas $\phi_{ik}$
  - DYF.eta: N x nocc.max x nb.chains lines and nb.etas columns
  - ind.ieta: lines with non-zero elements
- note: here done with eta but same idea could apply to phi (eg phiMc in the e-step)
- note: done directly as the mean over the nb.chains, not sure that in the code we don't need to keep the nb.chains separate during the E-step and only average them in the M-step

```{r etaStructure, results='hide', echo=FALSE, warnings=FALSE}
# Add to Dargs as re-used many times
nocc.max <- max(smx.data@var.N[[2]]) # max nb of occasions (to create a square matrix DYF.eta, with 0s for missing occasions)

# indices of lines with occasions present in the dataset
iocc<-rep(smx.data@var.N[[2]],each=nocc.max)
iall<-rep(1:nocc.max,Dargs$NM)
ind.ieta<-which(iall <= iocc)

# square matrix with (nocc*N*nb.chains) lines and nb.etas columns
DYF.eta <- matrix(0,nrow=Dargs$NM*nocc.max, ncol=nb.etas)

# generate random effects for all subjects for nocc.max occasions => etaMc has size Dargs$NM x (nb.etas*nocc.max)
etaMc<-matrix(rnorm(Dargs$NM*nb.etas.sim),ncol=nb.etas.sim)%*%rmat.omega
# transform etaMc to stack the occasions for each subject => size (Dargs$NM*nocc.max) x nb.etas
etaMc.occ <- matrix(etaMc, ncol=nb.etas, byrow=T)
DYF.eta[ind.ieta,] <- etaMc.occ[ind.ieta,]

# rescale the matrix into (nocc.max*nb.chains) rows x (nb.eta*N) subjects
# now we have all phi1, then all phi2, etc...
DYF.eta.tr<-matrix(DYF.eta,nrow=nocc.max*Dargs$nb.chains)

# compute the mean of each column, over only the available occasions (=mean(phi_ik) over the k occasions and the nb.chains chains)
inunits <- rep(smx.data@var.N[[2]],nb.etas) # nb of occasions for each subjects, repeated nb.etas times
mphi <- colSums(DYF.eta.tr)/Dargs$nb.chains/inunits

# If needed, we can put these back into a matrix N*nb.chains x nb.etas
mean.etaM <- matrix(mphi,ncol=nb.etas)

```

#### Computing likelihood for parameters (U.eta) 
  
- **ToDo**
  - figure out formula to replace 
        U.eta<-0.5*rowSums(etaM*(etaM%*%somega))
    - here etaM=$\phi_{ik}$
  - can we compare directly to extended matrix (but not invertible) ? probably not
  - do we need to compute $\phi_i$ and $\phi_{ik}$ and sum $\phi_i \; \Omega_1^{-1} \; \phi_i^T$ and $\sum_k \phi_{ik} \; \Omega_2^{-1} \; \phi_{ik}^T$ ? (probably)
  - here we also need to sum only on the levels of variability for each subject...
    - is there a clever way to do it ? (using indices, then using colSums and associated ni.occ)
- **maybe ToDo**
  - add nocc.max to the data object (currently computed in createVarlist)
  - add an element niocc to the levels of variability
    - here an obvious question: do we need to track the number of occasions or do we assume occasions are necessarily consecutive, 1, 2, 3 ?
  - add an index for the parameters corresponding to occasions (probably a list of size nocc.max)
    - maybe also need an equivalent of DYF for eta ? TBD

```{r simulPhi, results='hide', echo=FALSE, warnings=FALSE}
# Computing U.eta
etaM <- etaMc

if(FALSE) {
  # for the same reason, can't invert omega.eta
  somega<-solve(omega.eta)
  U.eta<-0.5*rowSums(etaM*(etaM%*%somega))
}

```

### Other

- *TBD: extension to more than 2 levels of variability (future)*

## Case-studies (isolated)

Matching data and object

- ToDo:
  - create functions to match data and object and check validity, updating with warning messages if need be

### Data and model without IOV, no covariates


```{r caseIIVnocov, results='hide', echo=FALSE, warnings=FALSE}

# Dargs from theodat.iiv (no covariates)
dlist <- createStructData(theodat.iiv, nb.chains=1)
Dargs.iiv <- dlist$Dargs
DYF.iiv <- dlist$DYF

# Margs from theomodel.iiv (IIV only)
mlist <- createStructModel(theomodel.iiv)
Margs.iiv <- mlist$Margs

vlist <- createVarlist(theomodel.iiv, theodat.iiv)
Vargs.iiv <- vlist$Vargs

```

### Data and model with IOV, no covariates


```{r caseIOVnocov, results='hide', echo=FALSE, warnings=FALSE}

# Dargs from theodat.iov (no covariates)
dlist <- createStructData(theodat.iov, nb.chains=1)
Dargs.iov <- dlist$Dargs
DYF.iov <- dlist$DYF

# Margs from theomodel.iov (IIV + IOV)
mlist <- createStructModel(theomodel.iov)
Margs.iov <- mlist$Margs

vlist <- createVarlist(theomodel.iov, theodat.iov)
Vargs.iov <- vlist$Vargs

```



### Sufficient statistics

- ToDo: M2 Lille
  - sufficient statistics should need only 
    - individual parameters (simulated or from psinocov) 
    - associated model predictions ()
    - observations from Dargs$yobs
    - residual error parameters from Margs$error.model
  - IOV model without covariates (formulas in paper by Panhard and Samson)
  - base model (no covariates, no IOV)
    - formulas in Kuhn and Lavielle TBC
  - model with covariates
    - linear regression to solve mu+beta.Ci = phi_i - eta_i ? TBD
    - without IIV: see formula in previous saemix
    - with IOV: not sure where/what level we apply this (probably at each variability level, needs to reconstitute the associated systems)
   
```{r suffStats, results='hide', echo=FALSE, warnings=FALSE}
# input: phiM
# Vargs: omega matrices 

psiM.iov <- as.matrix(psinocov[,c(2:4)])
phiM.iov <- transformPar(psiM.iov,Margs.iov$invtransform)

```


