---
title: "Changing classes in saemix"
author: "Emmanuelle Comets"
date: "2025-07-10"
output:
  pdf_document: default
  html_document: default
---

# Objective

**Highlight changes in class structures and in entering the input between current version of saemix (3.x) and next version (4.0)**

- several new classes created as building blocks to model classes
    - added a log slot to most classes to track warning messages so the user can review what changes were made during the creation of the object
    - class SaemixOutcome
      - define outcomes in the model (=responses)
      - used to compute gpred
    - class SaemixParameter
      - define parameters of the model
    - class SaemixVarModel
      - define variability model
    - class SaemixPopModel
      - define population-level individual parameters at each level of variability, eg for IIV $\phi_{pop,i}$ 
      - linear model (($\phi_{pop,i}=\mu + \sum \beta.cov_i$)
    - class SaemixIndivModel
      - define individual parameters $\phi_{i}$ by adding random effects to $\phi_{pop,i}$)
      - includes the list of SaemixPopModel and SaemixVarModel objects
- SaemixData object
  - main changes 
    - added a slot to specify the type of the outcomes
    - added a slot for grouping levels (varlevel) 
    - added an associated covariate.varlevel to specify where the covariate model should enter
    - added a log slot to track warning messages so the user can review what changes were made during the creation of the object
  - additional changes
    - log all warning messages to print out or keep track of in the object
- **ToDo SaemixData Eco**
  - update legacy tests from testthat_saemixData-read.R
  - update legacy tests from testthat_saemixData-covariates.R
- **ToDo SaemixData Lillois**
  - check recording of error messages to make sure as much as possible is captured/tracked
  - update legacy creator so it still works with the new class (error messages in legacy tests)
  - add default behaviour to define name.group as varlevel[1] if given (potentially deprecate both name.group and name.occ)
  - program a function to extract the covariate matrices for the different levels in varlevel
    - input: the object
    - output: the object, with a slot holding the list of matrices with the first value of the covariate at each level
      - eg: for id, first value of the covariate in each subject (size: N x nb of covariates not varying across id)
      - for occ, first value of the covariate in each subject at each occasion (size Sum_i nocc_i x nb of covariates varying with occasion)
      - etc... if more levels of variability
      - add (and record) warning messages if the covariate changes value within a grouping level
- SaemixIndivModel (**new**)
  - split in two parts, using lists of the two new classes defining:
    - the individual population model ($\phi_{pop,i}=\mu + \sum \beta.cov_i$)
    - the variability model: structure of covariance-matrix $\Omega$
  - functions associated with the variability model
  - functions associated with the individual population model
    - see notes and testthat
- SaemixModel object (**complete overhaul**)
  - child class of SaemixIndivModel
    - inherits the list of SaemixPopModel and SaemixVarModel objects
    - adds list of outcome in the model
    - adds the model function (and the sim.model function for non-Gaussian models)
  - **main changes**
  - multiple responses through the ytype identifier
    - definition of outcome through their distribution
      - continuous: normal
      - non-Gaussian: different types
    - error models for continuous responses
      - built-in error models
      - user-defined error models
  - definition of parameters through their distribution à la MLX-TRAN
    - built-in distributions with automatically defined transformation h and inverse h-1
    - in theory (untested), possible to write user-defined distributions
    - from the list of parameters, derive the variability levels
- **ToDo SaemixModel Lillois**
  - function to match varlevel between a model and an object
    - input: SaemixData and SaemixModel objects
    - output: TBD (could be both objects updated, or could be an updated varlevel and covariate.varlevel list for two functions to make the changes in the SaemixData and SaemixModel objects)
  - reprogram functions associated to the model class
    - predict: priority high, easy once matching between data and model done
    - summary: priority moderate, very easy
    - checkInitialFixedEffects: priority high, moderate difficulty once matching between data and model done
- **ToDo SaemixModel Eco**
  - reprogram plot functions associated to the model class
    - difficulty: high (dispatch according to response type)
    - code architecture involved
- input functions (**saemixData()** and **saemixModel()**)
  - legacy options to enter the model as previously
  - some new features 
  - **ToDo someone**
    - create legacy function saemixModel.legacy()
    - test that previous models (no iov, single responses) give the same object (content) with the legacy and the new creator functions
- Fisher Information Matrix by linearisation
  - scope: **TBD**
  - **ToDo Lillois**
    1. equations for the FIM using h and gradient(h) (*thèse Sylvie Retout*)
      - function derivative of h added to the SaemixPar object, check if derivative of inverse transform function also needed
      - ToDo: write equations and implement them 
    2. equations for the FIM in the presence of IOV  (*thèse Caroline Bazzoli TBC, ou Anne Dubois*, *Florence* , *Kristofferson et al. 2015*)
      - biblio methods
      - write equations 
      - implementation
    3. question: is the expression generalisable to more variability levels ??
  - additionally
    - FIM for multiple responses but already implemented by Lucas (?)
- further developments needed (out of scope for Lillois)
  - Eco
    - decide: structure for results or use the xxxHat children classes
    - add log slot in SaemixObject objects
    - make sure all warning messages are captured in the log slots for both data and model objects
  - Fisher Information Matrix by stochastic approximation (*Delattre et Kuhn Computo 2023*)
    - will need the same transformation as for the linearisation approach
    - extended to multiple responses by Alexandra
    - who: most likely Lucas or could be an intern project
  - testing user-defined error models
  - user-defined distributions
  - additional variability levels

# Setup

- folders
  - set **workDir** to the working directory
  - needs the saemix root folder (here **/home/eco/work/saemix/saemixextension**)
  - data will be saved in folder **data40**
- libraries
  - need to install ggplot2 and its sister libraries
  - need packages testthat (tests)
  - saemix package functions are loaded, from two versions
    - current version: folder **R**
    - next version: folder **Rext**

```{r setup, include=FALSE}
# Setting working directory
require("knitr")
knitr::opts_chunk$set(echo = TRUE)

# Libraries
library(rlang)
library(ggplot2) 
library(RColorBrewer) 
library(gridExtra)
library(viridis)
library(testthat)

#library(dplyr)
# library(Cairo)

# Working directory
saemixDir<-"/home/eco/work/saemix/saemixextension"
workDir <-getwd()
setwd(workDir)
knitr::opts_chunk$set(root.dir = workDir)

# Load function generics (same for both versions)
progDir.current<-file.path(saemixDir,"R")
progDir.next<-file.path(saemixDir,"Rext")
source(file.path(progDir.current,"aaa_generics.R"))

# directory
datDir<-file.path(saemixDir,"data40")

```


# Simulation of IOV with theophylline PK

- Features
  - theophylline PK
  - covariates
    - sex, weight changing with subject
    - CRCL, comedication changing with occasion
  - design
    - N=100 subjects
    - 1 to 3 occasion per subject (random)
    - sampling times for each occasion: 0.25, 1.5, 3.35, 12 and 24 h (sparse design in Dubois et al.)
    - dose: D=4mg
    - time between two occasions: 4 weeks (assume washout so no drug left)
- Model
  - PK response: 1 cpt model with absorption

    $$ Cpred = \frac{D}{V} \frac{k_a}{(k_a-CL/V)} \left( e^{- CL/V t} - e^{-k_a t} \right)$$
  - log-normal parameter distribution
  - population parameter values (Dubois et al.)
    - ka=1.48 h-1
    - CL=0.04 L/h
    - V=0.48 L
  - variabilities
    - IIV 
      - 30% on all parameters
      - correlation between CL and V (0.7)
    - IOV
      - IOV on CL only with 20% IOV [Error: simulated without IOV]
    - different from Dubois, where 2 scenarios, no correlation and no covariate effect
      - low IIV: 0.2 on ka and CL, 0.1 on V; half of IIV for IOV
      - regular BSV: 0.5 on ka, V, CL and 0.15 on IOV
    - combined residual error model with a=0.1 and b=0.1 (low IIV scenario from Dubois)
  - covariate models
    - weight: allometry on V and CL
    - gender: V decreased by 30% in women
    - CRCL: correlation with CL ($\beta_{CRCL, CL}=0.5$)
    - comedication: increase in CL by 30\%
- Not included in the simulation
  - NA values in the covariates
  - LOQ values

### Read datasets

```{r readSimulatedTheoData}
# Dataset without covariates
theonocov <- read.csv(file=file.path(datDir,"theoSimul_nocov.csv"), header=T, sep=",")

# Dataset with covariates
theocov <- read.csv(file=file.path(datDir,"theoSimul_cov.csv"), header=T, sep=",")
```

# Saemix 

Changes in saemix from previous to new version of the package

## Data object

## Functions from the current saemix package

- expected behaviour
  - reading the dataframe should work
  - appropriate columns should be recognised, including occ and ytype
  - plot: seems correct
  - no variability levels 
  
- ToDo:
  - maybe, if: we continue with current version [currently not the plan]
    - add automatic recognition for occ and ytype (currently not read unless explicitely set by user)
    - add "|occ" in the formula
    - check nesting order (check occasion is within id)
    - dispatch individual plots
  
```{r currentFctsData, results='hide', echo=FALSE, warnings=FALSE}
# Reading functions from current package (3.x)
source(file.path(progDir.current,"SaemixData.R"))
source(file.path(progDir.current,"SaemixData-methods.R"))
source(file.path(progDir.current,"SaemixData-methods_covariates.R"))

# Reading data into an saemixData object
smxdata.current <- saemixData(name.data=file.path(datDir,"theoSimul_nocov.csv"), header=T,na=".", sep=",", name.group=c("id"),name.predictors=c("tim", "dose"),name.response=c("ysim"), name.occ = "occ", unit=list(x="hr",y="mg/L"), verbose=TRUE)
x<-smxdata.current
xdata <- smxdata.current@data

# testthat checks
expect_equal(sort(unique(xdata$ytype)),c(1))
expect_equal(sort(unique(xdata$occ)),c(1,2,3))
expect_length(unique(xdata$tim),5)
expect_length(unique(xdata$dose),1)
 
# Plot of the data (mixing all responses)
plot(x)

# Plot to check if the data looks all right (separating responses and occasions)

occ.labs <-paste("Occasion",1:3)
names(occ.labs) <- c("1", "2", "3")

ggplot(xdata, aes(x=tim, y=ysim, group=id, colour=as.factor(occ))) + geom_point() + geom_line(data = subset(xdata)) + xlab("Time (hr)") + ylab("Concentrations (mg/L)")  + scale_color_viridis("Occasion",discrete = TRUE) +  facet_wrap(.~occ, labeller = labeller(occ=occ.labs, .multi_line=FALSE))

```

### Functions from the new saemix package

- **legacy use**
  - *method*: using the function **saemixData()** as previously
    - objective: seamless transition in legacy mode
  - expected behaviour
    - no error message
    - reading the dataframe should work
    - appropriate columns should be recognised, including occ and ytype
  - **new**: slot outcome
    - outcome are automatically mapped
    - here we expect the first two outcome to be recognised as continuous, and the third as binary will be mapped to categorical
  - systematic check on unchanged slots in the object: **passed**
- **new definition**
  - *method*: same function but explicit definition of outcome (name and type)
  - expected behaviour
    - no error message
    - reading the dataframe should work
    - appropriate columns should be recognised, including occ and ytype
  - **new slot** outcome
    - now we should have named responses with their prespecified type
  - **new slot** varlevel
    - automatically filled in with iiv and iov from name.group and name.occ
    - alternate definition through varlevel
  - systematic check on unchanged slots in the object: **passed**
- ToDo:
  - units for each response
    - add automatic recognition for occ and ytype (currently not read unless explicitely set by user)
    - add "|occ" in the formula
    - check nesting order (check occasion is within id)
    - dispatch individual plots
  - note: see tests for the updated class in corresponding testthat files

```{r newFctsData, results='hide', echo=FALSE, warnings=FALSE}
source(file.path(progDir.next,"SaemixData.R"))
source(file.path(progDir.next,"SaemixData-methods.R"))

############### Legacy 
# Reading data into an saemixData object - same instruction as before
smxdata.legacy <- saemixData(name.data=file.path(datDir,"theoSimul_nocov.csv"), header=T,na=".", sep=",", name.group=c("id"),name.predictors=c("tim", "dose"),name.response=c("ysim"), name.occ = "occ", unit=list(x="hr",y="mg/L"), verbose=TRUE)
x<-smxdata.legacy

## new slots outcome and varlevel automatically filled in
expect_equal(smxdata.legacy@outcome,c(ysim="continuous"))
expect_equal(smxdata.legacy@varlevel,c("id","occ"))

## Should give the same plot
xdata <- smxdata.legacy@data
ggplot(xdata, aes(x=tim, y=ysim, group=id, colour=as.factor(occ))) + geom_point() + geom_line(data = subset(xdata)) + xlab("Time (hr)") + ylab("Concentrations (mg/L)")  + scale_color_viridis("Occasion",discrete = TRUE) +  facet_wrap(.~occ, labeller = labeller(occ=occ.labs, .multi_line=FALSE))

############### New definition 
# Reading data into an saemixData object - specifying outcome type

smxdata.next <-  saemixData(name.data=file.path(datDir,"theoSimul_nocov.csv"), header=T,na=".", sep=",", name.group=c("id"),name.predictors=c("tim", "dose"),name.response=c("ysim"), name.occ = "occ", unit=list(x="hr",y="mg/L"), verbose=TRUE)
x<-smxdata.next

expect_equal(smxdata.next@outcome,c(ysim="continuous"))
expect_equal(smxdata.next@varlevel,c("id","occ"))

## Should give the same plot
xdata <- smxdata.next@data
ggplot(xdata, aes(x=tim, y=ysim, group=id, colour=as.factor(occ))) + geom_point() + geom_line(data = subset(xdata)) + xlab("Time (hr)") + ylab("Concentrations (mg/L)")  + scale_color_viridis("Occasion",discrete = TRUE) +  facet_wrap(.~occ, labeller = labeller(occ=occ.labs, .multi_line=FALSE))

```

### Detailed changes to Data object


```{r compareDataObjects, results='hide', echo=FALSE, warnings=FALSE}
#########################################################################
# systematic checks between current and legacy version- identical objects
x<-smxdata.legacy
object.slots <- names(attributes(x))
for(islot in object.slots) {
  if(!(islot %in% c("outcome", "sep", "na", "header","class","log","varlevel", "covariate.varlevel"))) {
    expect_equal(smxdata.current[islot], smxdata.legacy[islot])
  }
}

# testthat checks
xdata <- smxdata.legacy@data
expect_equal(sort(unique(xdata$ytype)),c(1))
expect_equal(sort(unique(xdata$occ)),c(1,2,3))
expect_length(unique(xdata$tim),5)
expect_length(unique(xdata$dose),1)

#########################################################################
# systematic checks between legacy and next version - identical objects
x<-smxdata.next
object.slots <- names(attributes(x))
for(islot in object.slots) {
  if(!(islot %in% c("outcome", "sep", "na", "header","class","units"))) {
    expect_equal(smxdata.next[islot], smxdata.legacy[islot])
  }
}

# testthat checks
xdata <- smxdata.next@data
expect_equal(sort(unique(xdata$ytype)),c(1))
expect_equal(sort(unique(xdata$occ)),c(1,2,3))
expect_length(unique(xdata$tim),5)
expect_length(unique(xdata$dose),1)
expect_equal(smxdata.next@outcome,c(ysim="continuous"))
expect_equal(smxdata.next@units$x,c("hr"))
expect_equal(smxdata.next@units$y,c("mg/L"))

```

## Model object

- model function used to estimate parameters
  - same arguments psi, id, xidep
  - ytype is a recognised column of xidep so can be used inside the code
- ToDo
  - add occ to xidep if not done yet
  - ToDo: add an option to use a column twice (eg occ could be used both as an indicator of variability level and as a predictor if there is a systematic change we want in the model) ? (another option=simply duplicate the column in the dataset with another name, which is the Monolix standard)
- TBD
  - here occ is also used (but need to make sure it is passed on in the code)
  - do we need occ in the model or should it also be passed 'within' psi ? (need to think; should be passed within psi, but could be passed by the user as a predictor as well)


```{r modelFunction, results='hide', echo=FALSE, warnings=FALSE}
# model function
theo1cpt <- function(psi,id,xidep) {
  tim<-xidep[,1]
  dose<-xidep[,2]
  ka<-psi[id,1]
  V<-psi[id,2]
  CL<-psi[id,3]
  ypred <- dose*ka/V/(ka-CL/V)*(exp(-CL/V*tim) - exp(-ka*tim))
  return(ypred)
}
```

## Functions from the current saemix package

- expected behaviour
  - only IIV taken into account
  - distributions, correlations between parameters handled
  - multiple responses can already be modelled through ytype
- limits for the current package
  - only IIV, no IOV
    - also of course no covariate model for IOV parameters
  - error model: only one taken into account
- ToDo:
  - vector of residual error models
  - maybe, if: we continue with current version [currently not the plan]
    - dispatch individual plots

```{r currentFctsModel, results='hide', echo=FALSE, warnings=FALSE}
# Reading functions from current package (3.x)
source(file.path(progDir.current,"SaemixModel.R"))

# Initial parameters & covariance matrix structure (IIV only)
psi0.mat <- as.matrix(data.frame(ka=1.5, V=0.5, CL=0.04))
varmat <- diag(1,nrow=dim(psi0.mat)[2])
varmat[2,3]<-varmat[3,2]<-1

smxmodel.current<-saemixModel(model=theo1cpt, modeltype=c("structural"),
  description="Theophylline PK model without IOV", 
  psi0=psi0.mat,transform.par=rep(1,3),covariance.model=varmat,error.model=c("combined"))

```

### Functions from the new saemix package

- methods for model object now in a separate file
- **legacy use**
  - *method*: using the **saemixModel()** function as previously
    - objective: seamless transition in legacy mode for models previously handled
  - expected behaviour
    - no error message
  - **new**: 
- **new definition without IOV**
  - *method*: same function but explicit definition of parameter distributions
  - expected behaviour
    - no error message
- **new definition with IOV**
  - *method*: include variability levels in the parameter definitions
- ToDo:
  - note: see tests for the updated class in corresponding testthat files
  - create legacy function to use the same instruction as before

```{r newFctsModel, results='hide', echo=FALSE, warnings=FALSE}
# Load subclasses for outcome, parameter, error model, variability model, individual, population and statistical models
source(file.path(progDir.next,"SaemixErrorModel.R"))
source(file.path(progDir.next,"SaemixOutcome.R"))
source(file.path(progDir.next,"SaemixParameter.R"))
source(file.path(progDir.next,"SaemixParameter-methods.R"))
source(file.path(progDir.next,"SaemixVarModel.R"))
source(file.path(progDir.next,"SaemixVarModel-methods.R"))
source(file.path(progDir.next,"SaemixPopModel.R"))
source(file.path(progDir.next,"SaemixIndivModel.R"))
source(file.path(progDir.next,"SaemixModel.R"))
source(file.path(progDir.next,"SaemixModel-methods.R"))

############### Legacy 
# Reading data into an saemixModel object - same instruction as before
smxmodel.legacy<-try(saemixModel.legacy(model=theo1cpt, modeltype=c("structural"),
  description="Theophylline PK model without IOV", 
  psi0=psi0.mat,transform.par=rep(1,3),covariance.model=varmat,error.model=c("combined")))
if(is(smxmodel.legacy,"try-error")) cat("ToDo: code a legacy function to recognise the old instructions\n")

############### New definition 
# Reading data into an saemixModel object - specifying parameter distributions

## Lazy way (IIV only)
smxmodel.lazy <- saemixModel(model=theo1cpt, parameters=c(ka=1.5, V=0.5, CL=0.04))

## New way, only IIV (like lazy)
smxmodel.noIOV <- saemixModel(model=theo1cpt, parameters=list(ka=saemixParam(mu.init=1.5), V=saemixParam(mu.init=0.5), CL=saemixParam(mu.init=0.04)))

############### New definition, with IOV in the model
## New way, full
param.base <- list(ka=saemixParam(mu.init=1.5),
                vd=saemixParam(mu.init=0.5, sd.init=0.7,varlevel=c("iiv","iov")),
                cl=saemixParam(name="cl",mu.init=0.04, varlevel=c("iiv","iov"), sd.init=c(0.6,0.3), corr = list(iiv=c("vd")))
                )

param.cov <- list(ka=saemixParam(mu.init=1.5),
                vd=saemixParam(mu.init=0.5, sd.init=0.7, covariate="wt", covariate.init=c(1), covariate.estim=c("fixed")),  
                cl=saemixParam(name="cl",mu.init=2, varlevel=c("iiv","iov"), sd.init=c(0.6,0.3), corr = list(iiv=c("vd")), covariate=c("wt","sex","age"), covariate.init=c(0.75,0,0), covariate.estim=c("fixed","estimated","estimated"), corr.init=list(iiv=c(-0.5,0.7), iov=0.7), covariate.varlevel=c("iiv","iiv","iov"))
                )
# Model without covariates
smxmodel.next <- saemixModel(model=theo1cpt, parameters=param.base)
cat("Variance structures:\n")
print(smxmodel.next@varmodel)

cat("Parameter structures:\n")
print(smxmodel.next@popmodel)

cat("Parameters mu & beta in the model\n")
showall(smxmodel.next@popmodel)
 
# Model with covariates
smxmodel.next.cov <- saemixModel(model=theo1cpt, parameters=param.cov)
cat("Variance structures:\n")
print(smxmodel.next.cov@varmodel)

cat("Parameter structures:\n")
print(smxmodel.next.cov@popmodel)

cat("Parameters mu & beta in the model\n")
showall(smxmodel.next.cov@popmodel)

```

### Detailed changes to Model object - ToDo

- **ToDo** code a legacy model function saemixModel.legacy() for users who wish to use their old code

```{r compareModelObjects, results='hide', echo=FALSE, warnings=FALSE}
#########################################################################
# systematic checks between next and legacy version- identical objects
if(!is(smxmodel.legacy,"try-error")) {
  object.slots <- names(attributes(x))
  for(islot in object.slots) {
    expect_equal(smxmodel.next[islot], smxdata.legacy[islot])
  }
}

# testthat checks

```

### Additional features (at some point)

- NA
- LOQ
