---
title: "Test saemix 3.0"
author: "Emmanuelle"
date: "05/10/2020"
output:
  pdf_document: default
  html_document: default
---

# Setup

- set up work directories
- two versions toggled by testMode
  - if testMode is FALSE, load the functions in R
  - if testMode is TRUE, load the library in a dev_mode environment

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Directories
saemixDir<-"/home/eco/work/saemix/saemixextension"
progDir<-file.path(saemixDir,"R")
datDir<-file.path(saemixDir,"data")
ecoDir<-file.path(saemixDir,"testeco")
belDir<-file.path(saemixDir,"testbelhal")

# Libraries
library(ggplot2)
library(MASS)

testMode<-TRUE

```

# Testing library

```{r testLibrary, include=FALSE}
# Loading library
if(testMode) {
  dev_mode()
  install.packages("saemix3.0.tar.gz",repos=NULL)
  library(saemix)
}
```

## Housekeeping and help functions

```{r}
cat("Check load message for date and author to contact \n")

# Library
if(testMode) {
  ?saemix
}
  
```


## Continuous response models

### Theophylline

```{r theo}
data(theo.saemix)

#Plotting the theophylline data
plot(Concentration~Time,data=theo.saemix,xlab="Time after dose (hr)",
ylab="Theophylline concentration (mg/L)")

saemix.data<-saemixData(name.data=theo.saemix,header=TRUE,sep=" ",na=NA, 
  name.group=c("Id"),name.predictors=c("Dose","Time"),
  name.response=c("Concentration"),name.covariates=c("Weight","Sex"),
  units=list(x="hr",y="mg/L",covariates=c("kg","-")), name.X="Time")
  model1cpt<-function(psi,id,xidep) { 
    dose<-xidep[,1]
    tim<-xidep[,2]  
    ka<-psi[id,1]
    V<-psi[id,2]
    CL<-psi[id,3]
    k<-CL/V
    ypred<-dose*ka/(V*(ka-k))*(exp(-k*tim)-exp(-ka*tim))
    return(ypred)
    }
# Default model, no covariate
saemix.model<-saemixModel(model=model1cpt,
       description="One-compartment model with first-order absorption",
       psi0=matrix(c(1.,20,0.5,0.1,0,-0.01),ncol=3,byrow=TRUE, 
       dimnames=list(NULL, c("ka","V","CL"))),transform.par=c(1,1,1))
 # Note: remove the options save=FALSE and save.graphs=FALSE 
 # to save the results and graphs
saemix.options<-list(seed=632545,save=FALSE,save.graphs=FALSE, displayProgress=FALSE)

saemix.fit<-saemix(saemix.model,saemix.data,saemix.options)
plot(saemix.fit)

# Model with covariates
saemix.model<-saemixModel(model=model1cpt,
                          description="One-compartment model with first-order absorption",
                          psi0=matrix(c(1.,20,0.5,0.1,0,-0.01),ncol=3,byrow=TRUE, 
                                      dimnames=list(NULL, c("ka","V","CL"))),transform.par=c(1,1,1), 
                          covariate.model=matrix(c(0,0,1,0,0,0),ncol=3,byrow=TRUE),fixed.estim=c(1,1,1),
                          covariance.model=matrix(c(1,0,0,0,1,1,0,1,1),ncol=3,byrow=TRUE),
                          omega.init=matrix(c(1,0,0,0,1,0,0,0,1),ncol=3,byrow=TRUE),error.model="combined")

saemix.options<-list(seed=39546,save=FALSE,save.graphs=FALSE, displayProgress=FALSE)
saemix.fit<-saemix(saemix.model,saemix.data,saemix.options)

```

### Simulated PD

```{r PD1}
  data(PD1.saemix)
  data(PD2.saemix)

saemix.data<-saemixData(name.data=PD1.saemix,name.group=c("subject"),
      name.predictors=c("dose"),name.response=c("response"),
      name.covariates=c("gender"), units=list(x="mg",y="-",covariates=c("-")))

modelemax<-function(psi,id,xidep) {
# input:
#   psi : matrix of parameters (3 columns, E0, Emax, EC50)
#   id : vector of indices 
#   xidep : dependent variables (same nb of rows as length of id)
# returns:
#   a vector of predictions of length equal to length of id
  dose<-xidep[,1]
  e0<-psi[id,1]
  emax<-psi[id,2]
  e50<-psi[id,3]
  f<-e0+emax*dose/(e50+dose)
  return(f)
}

# Plotting the data
plot(saemix.data,main="Simulated data PD1")
# Compare models with and without covariates with LL by Importance Sampling
model1<-saemixModel(model=modelemax,description="Emax growth model", 
       psi0=matrix(c(20,300,20,0,0,0),ncol=3,byrow=TRUE,dimnames=list(NULL,
       c("E0","Emax","EC50"))), transform.par=c(1,1,1),
       covariate.model=matrix(c(0,0,0), ncol=3,byrow=TRUE),fixed.estim=c(1,1,1))

model2<-saemixModel(model=modelemax,description="Emax growth model", 
       psi0=matrix(c(20,300,20,0,0,0),ncol=3,byrow=TRUE,dimnames=list(NULL, 
       c("E0","Emax","EC50"))), transform.par=c(1,1,1),
       covariate.model=matrix(c(0,0,1), ncol=3,byrow=TRUE),fixed.estim=c(1,1,1))

# SE not computed as not needed for the test
saemix.options<-list(algorithms=c(0,1,1),nb.chains=3,seed=765754, 
       nbiter.saemix=c(500,300),save=FALSE,save.graphs=FALSE)

fit1<-saemix(model1,saemix.data,saemix.options)
fit2<-saemix(model2,saemix.data,saemix.options)
wstat<-(-2)*(fit1["results"]["ll.is"]-fit2["results"]["ll.is"])

cat("LRT test for covariate effect on EC50: p-value=",1-pchisq(wstat,1),"\n")

```


### Oxford boys

```{r oxfordBoys}
data(oxboys.saemix)

saemix.data<-saemixData(name.data=oxboys.saemix,
      name.group=c("Subject"),name.predictors=c("age"),name.response=c("height"),
      units=list(x="yr",y="cm"))

# plot the data
plot(saemix.data)

growth.linear<-function(psi,id,xidep) {
  x<-xidep[,1]
  base<-psi[id,1]
  slope<-psi[id,2]
  f<-base+slope*x
  return(f)
}
saemix.model<-saemixModel(model=growth.linear,description="Linear model",
      psi0=matrix(c(140,1),ncol=2,byrow=TRUE,dimnames=list(NULL,c("base","slope"))),
      transform.par=c(1,0),covariance.model=matrix(c(1,1,1,1),ncol=2,byrow=TRUE), 
      error.model="constant")

saemix.options<-list(algorithms=c(1,1,1),nb.chains=1,seed=201004,
      save=FALSE,save.graphs=FALSE)
saemix.fit<-saemix(saemix.model,saemix.data,saemix.options)
```

### Cow

```{r cow}
data(cow.saemix)

saemix.data<-saemixData(name.data=cow.saemix,name.group=c("cow"), 
      name.predictors=c("time"),name.response=c("weight"), 
      name.covariates=c("birthyear","twin","birthrank"), 
      units=list(x="days",y="kg",covariates=c("yr","-","-")))

growthcow<-function(psi,id,xidep) {
  x<-xidep[,1]
  a<-psi[id,1]
  b<-psi[id,2]
  k<-psi[id,3]
  f<-a*(1-b*exp(-k*x))
  return(f)
}
saemix.model<-saemixModel(model=growthcow,
      description="Exponential growth model", 
      psi0=matrix(c(700,0.9,0.02,0,0,0),ncol=3,byrow=TRUE, 
        dimnames=list(NULL,c("A","B","k"))),transform.par=c(1,1,1),fixed.estim=c(1,1,1), 
      covariate.model=matrix(c(0,0,0),ncol=3,byrow=TRUE), 
      covariance.model=matrix(c(1,0,0,0,1,0,0,0,1),ncol=3,byrow=TRUE), 
      omega.init=matrix(c(1,0,0,0,1,0,0,0,1),ncol=3,byrow=TRUE),error.model="constant")

saemix.options<-list(algorithms=c(1,1,1),nb.chains=1,nbiter.saemix=c(200,100), 
             seed=4526,save=FALSE,save.graphs=FALSE)

# Plotting the data
plot(saemix.data,xlab="Time (day)",ylab="Weight of the cow (kg)")
saemix.fit<-saemix(saemix.model,saemix.data,saemix.options)
```


### Wheat yield

```{r yield}
data(yield.saemix)
saemix.data<-saemixData(name.data=yield.saemix,name.group=c("site"),
      name.predictors=c("dose"),name.response=c("yield"),
      name.covariates=c("soil.nitrogen"),units=list(x="kg/ha",y="t/ha",covariates=c("kg/ha")))

#  Model: linear + plateau
yield.LP<-function(psi,id,xidep) {
  x<-xidep[,1]
  ymax<-psi[id,1]
  xmax<-psi[id,2]
  slope<-psi[id,3]
  f<-ymax+slope*(x-xmax)
  #'  cat(length(f),"  ",length(ymax),"\n")
  f[x>xmax]<-ymax[x>xmax]
  return(f)
}
saemix.model<-saemixModel(model=yield.LP,description="Linear plus plateau model",   
        psi0=matrix(c(8,100,0.2,0,0,0),ncol=3,byrow=TRUE,dimnames=list(NULL,   
            c("Ymax","Xmax","slope"))),covariate.model=matrix(c(0,0,0),ncol=3,byrow=TRUE), 
        transform.par=c(0,0,0),covariance.model=matrix(c(1,0,0,0,1,0,0,0,1),ncol=3, 
            byrow=TRUE),error.model="constant")

saemix.options<-list(algorithms=c(1,1,1),nb.chains=1,seed=666, 
       save=FALSE,save.graphs=FALSE)

# Plotting the data
plot(saemix.data,xlab="Fertiliser dose (kg/ha)", ylab="Wheat yield (t/ha)")

saemix.fit<-saemix(saemix.model,saemix.data,saemix.options)

# Comparing the likelihoods obtained by linearisation and importance sampling 
# to the likelihood obtained by Gaussian Quadrature
saemix.fit<-llgq.saemix(saemix.fit)
{
   cat("LL by Importance sampling, LL_IS=",saemix.fit["results"]["ll.is"],"\n")
   cat("LL by linearisation, LL_lin=",saemix.fit["results"]["ll.lin"],"\n")
   cat("LL by Gaussian Quadrature, LL_GQ=",saemix.fit["results"]["ll.gq"],"\n")
}

# Testing for an effect of covariate soil.nitrogen on Xmax
saemix.model2<-saemixModel(model=yield.LP,description="Linear plus plateau model", 
         psi0=matrix(c(8,100,0.2,0,0,0),ncol=3,byrow=TRUE,dimnames=list(NULL, 
            c("Ymax","Xmax","slope"))),covariate.model=matrix(c(0,1,0),ncol=3,byrow=TRUE), 
         transform.par=c(0,0,0),covariance.model=matrix(c(1,0,0,0,1,0,0,0,1),ncol=3, 
             byrow=TRUE),error.model="constant")

saemix.fit2<-saemix(saemix.model2,saemix.data,saemix.options)
# BIC for the two models
{
  cat("Model without covariate, BIC=",saemix.fit["results"]["bic.is"],"\n")
  cat("Model with covariate, BIC=",saemix.fit2["results"]["bic.is"],"\n")
  pval<-1-pchisq(-2*saemix.fit["results"]["ll.is"]+2*saemix.fit2["results"]["ll.is"],1)
  cat("        LRT: p=",pval,"\n")
}

```

## Discrete data model


### Binary response model

Toenail data

```{r binary}
data(toenail.saemix) 

saemix.data<-saemixData(name.data=toenail.saemix,name.group=c("id"),name.predictors=c("time","y"), name.response="y",
                        name.covariates=c("treatment"),name.X=c("time"))

binary.model<-function(psi,id,xidep) {
  tim<-xidep[,1]
  y<-xidep[,2]
  inter<-psi[id,1]
  slope<-psi[id,2]
  logit<-inter+slope*tim
  pevent<-exp(logit)/(1+exp(logit))
  logpdf<-rep(0,length(tim))
  P.obs = (y==0)*(1-pevent)+(y==1)*pevent
  logpdf <- log(P.obs)
  return(logpdf)
}

saemix.model<-saemixModel(model=binary.model,description="Binary model",
                          modeltype="likelihood",
                          psi0=matrix(c(0,-.5,0,0.5),ncol=2,byrow=TRUE,dimnames=list(NULL,c("theta1","theta2"))),
                          transform.par=c(0,0), covariate.model=c(0,1),covariance.model=matrix(c(1,0,0,1),ncol=2))

saemix.options<-list(seed=1234567,save=FALSE,save.graphs=FALSE, displayProgress=FALSE, nb.chains=10, fim=FALSE)

binary.fit<-saemix(saemix.model,saemix.data,saemix.options)
```

### Categorical response model

- Knee pain after 3, 7 and 10 days of treatment compared to baseline (time=0)
  - longitudinal ordinal model with 5 categories
- Comparing the 3 covariate models - model with Age on alp1 and treatment on beta best

```{r knee}
data(knee.saemix)

saemix.data<-saemixData(name.data=knee.saemix,name.group=c("id"),
                        name.predictors=c("y", "time"), name.X=c("time"),
                        name.covariates = c("Age","Sex","treatment"))

ordinal.model<-function(psi,id,xidep) {
  y<-xidep[,1]
  time<-xidep[,2]
  alp1<-psi[id,1]
  alp2<-psi[id,2]
  alp3<-psi[id,3]
  alp4<-psi[id,4]
  beta<-psi[id,5]
  
  logit1<-alp1 + beta*time
  logit2<-logit1+alp2
  logit3<-logit2+alp3
  logit4<-logit3+alp4
  pge1<-exp(logit1)/(1+exp(logit1))
  pge2<-exp(logit2)/(1+exp(logit2))
  pge3<-exp(logit3)/(1+exp(logit3))
  pge4<-exp(logit4)/(1+exp(logit4))
  logpdf<-rep(0,length(y))
  P.obs = (y==1)*pge1+(y==2)*(pge2 - pge1)+(y==3)*(pge3 - pge2)+(y==4)*(pge4 - pge3)+(y==5)*(1 - pge4)
  logpdf <- log(P.obs)
  
  return(logpdf)
}
covmodel3<-covmodel2<-covmodel1<-matrix(data=0,ncol=5,nrow=3)
covmodel1[1:2,1]<-1
covmodel1[,5]<-1
covmodel2[1,1]<-covmodel2[3,5]<-1
covmodel2<-covmodel<-matrix(data=0,ncol=5,nrow=3)
covmodel3[1,1]<-1

saemix.model<-saemixModel(model=ordinal.model,description="Ordinal categorical model",modeltype="likelihood",
                          psi0=matrix(c(0,0.2, 0.6, 3, 0.2),ncol=5,byrow=TRUE,dimnames=list(NULL,c("alp1","alp2","alp3","alp4","beta"))),
                          transform.par=c(0,1,1,1,1),omega.init=diag(rep(1,5)), covariance.model = diag(c(1,0,0,0,1)))

saemix.model.cov1<-saemixModel(model=ordinal.model,description="Ordinal categorical model",modeltype="likelihood",
                          psi0=matrix(c(0,0.2, 0.6, 3, 0.2),ncol=5,byrow=TRUE,dimnames=list(NULL,c("alp1","alp2","alp3","alp4","beta"))),
                          transform.par=c(0,1,1,1,1),omega.init=diag(rep(1,5)), covariance.model = diag(c(1,0,0,0,1)),
                          covariate.model = covmodel)
saemix.model.cov2<-saemixModel(model=ordinal.model,description="Ordinal categorical model",modeltype="likelihood",
                              psi0=matrix(c(0,0.2, 0.6, 3, 0.2),ncol=5,byrow=TRUE,dimnames=list(NULL,c("alp1","alp2","alp3","alp4","beta"))),
                              transform.par=c(0,1,1,1,1),omega.init=diag(rep(1,5)), covariance.model = diag(c(1,0,0,0,1)),
                              covariate.model = covmodel2)
saemix.model.cov3<-saemixModel(model=ordinal.model,description="Ordinal categorical model",modeltype="likelihood",
                               psi0=matrix(c(0,0.2, 0.6, 3, 0.2),ncol=5,byrow=TRUE,dimnames=list(NULL,c("alp1","alp2","alp3","alp4","beta"))),
                               transform.par=c(0,1,1,1,1),omega.init=diag(rep(1,5)), covariance.model = diag(c(1,0,0,0,1)),
                               covariate.model = covmodel3)

saemix.options<-list(seed=632545,save=FALSE,save.graphs=FALSE, nb.chains=5)

ord.fit<-saemix(saemix.model,saemix.data,saemix.options)
ord.fit.cov1<-saemix(saemix.model.cov1,saemix.data,saemix.options)
ord.fit.cov2<-saemix(saemix.model.cov2,saemix.data,saemix.options)
ord.fit.cov3<-saemix(saemix.model.cov3,saemix.data,saemix.options)

# Comparing the 3 covariate models - model with Age on alp1 and treatment on beta best
compare.saemix(ord.fit.cov1,ord.fit.cov2, ord.fit.cov3)

```


### Count data model

**TODO**

##  Time-to-event

### TTE model

```{r lungTTE}
data(lung.saemix)

saemix.data<-saemixData(name.data=lung.saemix,header=TRUE,name.group=c("id"),
      name.predictors=c("time","status","cens"),name.response=c("status"),
      name.covariates=c("age", "sex", "ph.ecog", "ph.karno", "pat.karno", "wt.loss","meal.cal"),
      units=list(x="days",y="",covariates=c("yr","","-","%","%","cal","pounds")))

weibulltte.model<-function(psi,id,xidep) {
  T<-xidep[,1]
  y<-xidep[,2] # events (1=event, 0=no event)
  cens<-which(xidep[,3]==1) # censoring times (subject specific)
  init <- which(T==0)
  lambda <- psi[id,1] # Parameters of the Weibull model
  beta <- psi[id,2]
  Nj <- length(T)
  
  ind <- setdiff(1:Nj, append(init,cens)) # indices of events
  hazard <- (beta/lambda)*(T/lambda)^(beta-1) # ln(H')
  H <- (T/lambda)^beta # ln(H)
  logpdf <- rep(0,Nj) # ln(l(T=0))=0
  logpdf[cens] <- -H[cens] + H[cens-1] # ln(l(T=censoring time))
  logpdf[ind] <- -H[ind] + H[ind-1] + log(hazard[ind]) # ln(l(T=event time))
  return(logpdf)
}

saemix.model<-saemixModel(model=weibulltte.model,description="time model",modeltype="likelihood",
  psi0=matrix(c(1,2),ncol=2,byrow=TRUE,dimnames=list(NULL,  c("lambda","beta"))),
  transform.par=c(1,1),covariance.model=matrix(c(1,0,0,0),ncol=2, byrow=TRUE))
saemix.options<-list(seed=632545,save=FALSE,save.graphs=FALSE, displayProgress=FALSE)
tte.fit<-saemix(saemix.model,saemix.data,saemix.options)
plot(tte.fit, plot.type="convergence")

ypred<-predict(tte.fit)

```


### RTTE

**TODO**

## Exiting dev mode

```{r testLibraryEnd, include=FALSE}
if(testMode) 
  dev_mode()
```
