---
title: "Test saemix 3.0"
author: "Emmanuelle"
date: "05/10/2020"
output:
  pdf_document: default
  html_document: default
---

# Setup

- set up work directories
- two versions toggled by testMode
  - if testMode is FALSE, load the functions in R
  - if testMode is TRUE, load the library in a dev_mode environment

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Directories
saemixDir<-"/home/eco/work/saemix/saemixextension"
progDir<-file.path(saemixDir,"R")
datDir<-file.path(saemixDir,"data")
ecoDir<-file.path(saemixDir,"testeco")
belDir<-file.path(saemixDir,"testbelhal")

# Libraries
library(ggplot2)
library(MASS)

testMode<-FALSE

```

# Testing library

```{r testLibrary, include=FALSE}
# Sourcing saemix functions
if(!testMode) {
  source(file.path(progDir,"aaa_generics.R"))
  #source(file.path(progDir,"global.R"))
  source(file.path(progDir,"SaemixData.R"))
  source(file.path(progDir,"SaemixRes.R"))
  source(file.path(progDir,"SaemixModel.R"))
  source(file.path(progDir,"SaemixObject.R"))
  source(file.path(progDir,"main.R"))
  source(file.path(progDir,"func_aux.R"))
  source(file.path(progDir,"main_initialiseMainAlgo.R"))
  source(file.path(progDir,"main_estep.R"))
  source(file.path(progDir,"main_mstep.R"))
  source(file.path(progDir,"func_FIM.R"))
  source(file.path(progDir,"func_plots.R"))
  source(file.path(progDir,"func_distcond.R"))
  source(file.path(progDir,"func_simulations.R"))
  source(file.path(progDir,"compute_LL.R"))
  source(file.path(progDir,"func_estimParam.R"))
}
# Loading library
if(testMode) {
  dev_mode()
  install.packages("saemix3.0.tar.gz",repos=NULL)
  library(saemix)
}
```

## Housekeeping and help functions

```{r}
cat("Check load message for date and author to contact \n")

# Library
if(testMode) {
  ?saemix
}
  
```


## Continuous response model

## Discrete data model

### Binary response model

### Categorical response model

### Count data model

##  Time-to-event

### TTE model

TTE data simulated according to a Weibull model, hazard defined by shape ($\beta$) and scale ($\lambda$) as:
$$ h(t) = \frac{\beta}{\lambda} \left( \frac{t}{\lambda} \right) ^{\beta-1}$$

```{r TTEsimul}
# Simulating TTE data
set.seed(12345)

nsuj<-50
xtim<-c(0)
tte.data<-data.frame(id=rep(1:nsuj,each=length(xtim)),time=rep(xtim,nsuj))
psiM<-data.frame(lambda=seq(1.6,2,length.out=length(unique(tte.data$id))),beta = 2)

simul.tte<-function(psi,id,xidep) {
  T<-xidep
  N <- nrow(psi)
  Nj <- length(T)
  censoringtime = 3
  lambda <- psi[id,1]
  beta <- psi[id,2]
  obs <-rep(0,length(T))
  for (i in (1:N)){
    obs[id==i] <- rweibull(n=length(id[id==i]), shape=beta[i], scale=lambda[i])
  } 
  obs[obs>censoringtime]<-censoringtime
  return(obs)
}

preds <- simul.tte(psiM, tte.data$id, tte.data[,c("time")])
tte.data$y<-0
tte.data$tlat<-preds
dat1<-tte.data[,c("id","time","y")]
dat2<-tte.data[,c("id","tlat","y")]
dat2$y<-as.integer(dat2$tlat>0 & dat2$tlat<3)
colnames(dat2)[2]<-"time"
tte.data<-rbind(dat1,dat2)
tte.data<-tte.data[order(tte.data$id, tte.data$time),]
tte.psiM<-psiM

# Simulate T from Weibull (check)
if(FALSE) { 
  lambda<-2
  beta<-2
  nsim<-5000
  # By hand
  q1<-runif(nsim)
  #  tevent<-lambda*exp(log(q1)/beta)
  tevent<-lambda*exp(log(-log(q1))/beta)
  tevent<-sort(tevent)
#  plot(tevent, exp(-(tevent/lambda)^beta))
  tevent2<-sort(rweibull(nsim, shape=beta, scale=lambda))
  plot(tevent, tevent2)
  abline(0,1)
   
}
```

```{r fitTTE}
saemix.data<-saemixData(name.data=tte.data, name.group=c("id"),
  name.predictors=c("time"), name.response="y")

tte.model<-function(psi,id,xidep) {
  T<-xidep[,1]
  N <- nrow(psi)
  Nj <- length(T)
  # censoringtime = 6
  censoringtime = max(T)
  lambda <- psi[id,1]
  beta <- psi[id,2]
  init <- which(T==0)
  cens <- which(T==censoringtime)
  ind <- setdiff(1:Nj, append(init,cens))
  hazard <- (beta/lambda)*(T/lambda)^(beta-1)
  H <- (T/lambda)^beta
  logpdf <- rep(0,Nj)
  logpdf[cens] <- -H[cens] + H[cens-1]
  logpdf[ind] <- -H[ind] + H[ind-1] + log(hazard[ind])
  return(logpdf)
}

saemix.model<-saemixModel(model=tte.model,description="time model",modeltype="likelihood",
  psi0=matrix(c(1,2),ncol=2,byrow=TRUE,dimnames=list(NULL,  c("lambda","beta"))),
  transform.par=c(1,1),covariance.model=matrix(c(1,0,0,1),ncol=2, byrow=TRUE))
saemix.model<-saemixModel(model=tte.model,description="time model",modeltype="likelihood",
  psi0=matrix(c(1,2),ncol=2,byrow=TRUE,dimnames=list(NULL,  c("lambda","beta"))),
  transform.par=c(1,1),covariance.model=matrix(c(1,0,0,0),ncol=2, byrow=TRUE))
saemix.options<-list(seed=632545,save=FALSE,save.graphs=FALSE, displayProgress=FALSE)
tte.fit<-saemix(saemix.model,saemix.data,saemix.options)
plot(tte.fit, plot.type="convergence")
```

### RTTE


```{r RTTEsimul}
# Simulating TTE data
set.seed(12345)

nsuj<-50
xtim<-c(0)
tte.data<-data.frame(id=rep(1:nsuj,each=length(xtim)),time=rep(xtim,nsuj))
psiM<-data.frame(lambda=seq(1.6,2,length.out=length(unique(tte.data$id))),beta = 2)

simul.tte<-function(psi,id,xidep) {
  T<-xidep
  N <- nrow(psi)
  Nj <- length(T)
  censoringtime = 3
  lambda <- psi[id,1]
  beta <- psi[id,2]
  obs <-rep(0,length(T))
  for (i in (1:N)){
    obs[id==i] <- rweibull(n=length(id[id==i]), shape=beta[i], scale=lambda[i])
  } 
  obs[obs>censoringtime]<-censoringtime
  return(obs)
}

preds <- simul.tte(psiM, tte.data$id, tte.data[,c("time")])
tte.data$y<-0
tte.data$tlat<-preds
dat1<-tte.data[,c("id","time","y")]
dat2<-tte.data[,c("id","tlat","y")]
dat2$y<-as.integer(dat2$tlat>0 & dat2$tlat<3)
colnames(dat2)[2]<-"time"
tte.data<-rbind(dat1,dat2)
tte.data<-tte.data[order(tte.data$id, tte.data$time),]
tte.psiM<-psiM

# Simulate T from Weibull (check)
if(FALSE) { 
  lambda<-2
  beta<-2
  nsim<-5000
  # By hand
  q1<-runif(nsim)
  #  tevent<-lambda*exp(log(q1)/beta)
  tevent<-lambda*exp(log(-log(q1))/beta)
  tevent<-sort(tevent)
#  plot(tevent, exp(-(tevent/lambda)^beta))
  tevent2<-sort(rweibull(nsim, shape=beta, scale=lambda))
  plot(tevent, tevent2)
  abline(0,1)
   
}
```

```{r fitRTTE}
saemix.data<-saemixData(name.data=tte.data, name.group=c("id"),
  name.predictors=c("time"), name.response="y")

tte.model<-function(psi,id,xidep) {
  T<-xidep[,1]
  N <- nrow(psi)
  Nj <- length(T)
  # censoringtime = 6
  censoringtime = max(T)
  lambda <- psi[id,1]
  beta <- psi[id,2]
  init <- which(T==0)
  cens <- which(T==censoringtime)
  ind <- setdiff(1:Nj, append(init,cens))
  hazard <- (beta/lambda)*(T/lambda)^(beta-1)
  H <- (T/lambda)^beta
  logpdf <- rep(0,Nj)
  logpdf[cens] <- -H[cens] + H[cens-1]
  logpdf[ind] <- -H[ind] + H[ind-1] + log(hazard[ind])
  return(logpdf)
}

saemix.model<-saemixModel(model=tte.model,description="time model",modeltype="likelihood",
  psi0=matrix(c(1,2),ncol=2,byrow=TRUE,dimnames=list(NULL,  c("lambda","beta"))),
  transform.par=c(1,1),covariance.model=matrix(c(1,0,0,1),ncol=2, byrow=TRUE))
saemix.model<-saemixModel(model=tte.model,description="time model",modeltype="likelihood",
  psi0=matrix(c(1,2),ncol=2,byrow=TRUE,dimnames=list(NULL,  c("lambda","beta"))),
  transform.par=c(1,1),covariance.model=matrix(c(1,0,0,0),ncol=2, byrow=TRUE))
saemix.options<-list(seed=632545,save=FALSE,save.graphs=FALSE, displayProgress=FALSE)
tte.fit<-saemix(saemix.model,saemix.data,saemix.options)
plot(tte.fit, plot.type="convergence")
```
## Exiting dev mode

```{r testLibraryEnd, include=FALSE}
if(testMode) 
  dev_mode()
```
